# Soil Spectroscopy for Digital Soil Mapping

*Alexandre Wadoux and Leonardo Ramirez-Lopez*

```{r setup, include=FALSE}
required_packages <- c(
  "terra", "ranger", "dplyr", "ggplot2",
  "prospectr", "matrixStats", "DiagrammeR",
  "DiagrammeRsvg", "rsvg"
)

dir.create("images/module4", recursive = TRUE, showWarnings = FALSE)

missing <- required_packages[!sapply(required_packages, requireNamespace, quietly = TRUE)]
if (length(missing) > 0) install.packages(missing)


library(terra)
library(ranger)
library(dplyr)
library(ggplot2)
library(prospectr)
library(matrixStats)
library(DiagrammeR)
library(DiagrammeRsvg)
library(rsvg)

knitr::opts_chunk$set(
  echo = TRUE,
  eval = FALSE,
  message = FALSE,
  warning = FALSE, 
  fig.align = 'center'
)
```

## Part I - Digital soil mapping with spectroscopy 

### Introduction 

Digital soil mapping (DSM) relies heavily on laboratory analyses to measure and characterize soil properties at sampled locations, which are then used as reference observations for training and validating spatial prediction models. Although highly accurate, conventional laboratory methods are slow and costly, creating a bottleneck that restricts timely data production for DSM. For example, conventional soil organic carbon reference methods, such as dry combustion and wet oxidation, require specialized infrastructure, trained personnel, and consumables, leading to long turnaround times and substantial per-sample costs [@wu2025assessing]. In this context, sensing technologies such as diffuse reflectance spectroscopy can complement laboratory analysis by reducing analytical costs and enabling soil property observations to be generated more rapidly and at greater spatial density. For these reasons, diffuse reflectance spectroscopy has gained substantial attention in the soil science community as a tool for enhancing soil characterization [@peng2025spectroscopic].

In soil applications, diffuse reflectance spectroscopy (DRS) is a maturing spectroscopic approach focused on extracting information from the interaction between soil samples and electromagnetic radiation. The chemical composition and physical structure of soil determine how electromagnetic radiation is absorbed and scattered. Consequently, soil composition and physical properties can be inferred by analyzing reflected energy. DRS sensors record reflected energy across multiple wavelengths, producing spectra that are typically converted from reflectance to absorbance. Soil property quantification with DRS relies on empirical calibration models, usually built with regression methods that link spectral measurements to laboratory reference data. Once suitable calibration models are available for multiple soil properties, they can be applied to a new soil spectrum to predict those properties simultaneously.

DRS methods used in soil science are commonly grouped into visible–near infrared (vis–NIR), shortwave infrared (SWIR), and mid-infrared (MIR) regions. The vis–NIR and SWIR regions are used in both laboratory and field applications, supported by the availability of portable instruments. These regions show a high sensitivity to soil organic matter, moisture, iron oxides, and texture-related features. MIR spectroscopy is more commonly conducted under laboratory conditions and often provides sharper absorption features associated with mineralogical and chemical soil attributes. In all these regions, spectral measurements of a soil sample can typically be acquired within seconds to a few minutes.

There is broad consensus that DRS complements rather than replaces soil laboratory analyses [@viscarra2022diffuse, @chen2023integrating], increasing analytical throughput while retaining the validity of reference procedures. For DSM, this means that spectroscopy can expand soil property datasets and improve mapping workflows while reducing cost and logistical constraints. This is true when the empirical models that translate spectra into soil property predictions have been properly calibrated and validated. It is important to note, however, that developing these models can still be laborious, time-consuming, and costly, a limitation that is rarely acknowledged in the soil spectroscopy literature. In this respect, pre-existing soil spectral libraries containing reference samples offer an opportunity to reduce calibration effort when developing new models for a specific area or application.


### Rationale 

The use of DRS-inferred soil properties offers a cost- and time-efficient pathway to substantially increase the number of observations available for DSM model calibration [@ramirez2019robust]. This directly supports the collection of spatially dense datasets at field to regional scales. By augmenting conventional laboratory datasets with larger volumes of DRS-inferred observations, DSM workflows can benefit from more input data, which may improve map accuracy at lower overall cost.

### Including Spectroscopic Predictions in Digital Soil Mapping

In DSM workflows that incorporate DRS, spectroscopically derived soil property estimates should be treated as model-based predictions with associated uncertainty, rather than as error-free measurements. Unlike conventional laboratory analyses, spectroscopy does not directly measure soil properties; it generates predictions from statistical models calibrated to capture relationships between spectral responses and reference measurements. Consequently, each DRS estimate is associated with uncertainty that is usually larger than that of conventional laboratory measurements. When DRS-inferred data are incorporated into DSM workflows, this additional source of uncertainty should therefore be explicitly acknowledged [@takoutsing2022accounting]. Treating DRS predictions as error-free observations may lead to biased spatial models, propagation of systematic spectroscopic errors, and overconfident DSM predictions [@van2022measurement].

Different methodological strategies have been proposed to propagate or account for uncertainty in soil property observations used in DSM, depending on the modeling framework adopted. In geostatistical mapping, observation-level uncertainty in soil property data has been incorporated into spatial covariance structures [@ramirez2019robust; @takoutsing2022accounting]. In machine-learning-based DSM, the literature remains more limited, although several studies have explored weighting schemes to reflect differences in data quality. For example, weights derived from DRS prediction uncertainty have been used in convolutional neural networks and random forest models by modifying the loss function during model calibration [@wadoux2019multi; @hengl2018random]. More recently, error-filtered machine learning frameworks have been proposed in which observation-level uncertainty is explicitly integrated into likelihood-based estimation, with improved predictive performance when observation errors are substantial [@van2022measurement].

A prerequisite for accounting for uncertainty in DRS predictions within DSM is that each observation includes an associated uncertainty estimate. In practice, DRS should provide not only a predicted soil property value, such as soil organic carbon content, but also an uncertainty metric, typically expressed as a prediction standard deviation or variance.

This is illustrated in Figure \@ref(fig:fig-uncertainty), which provides a conceptual representation of soil property observations obtained using conventional laboratory analysis and DRS. The laboratory measurement is shown as a narrow probability distribution centered on the reference value, reflecting relatively low analytical uncertainty. In contrast, the DRS-inferred value is represented by a wider probability distribution centered on the model prediction, reflecting larger prediction uncertainty. The dashed vertical lines denote the corresponding point estimates, while the horizontal arrows represent their standard deviations, $\sigma_{\mathrm{lab}}$ and $\sigma_{\mathrm{spectroscopy}}$. The displacement between the two distributions illustrates that spectroscopic predictions may differ systematically from reference measurements, introducing prediction bias in addition to random uncertainty.

DRS predictions of soil properties are generally less accurate than conventional laboratory measurements. However, their lower cost and higher analytical throughput can enable the acquisition of spatially denser datasets, which may improve the characterization of spatial variation and soil–environment relationships, provided that the associated prediction uncertainty is explicitly considered. In the practical section that follows, we illustrate this concept using an example of DSM implemented with a random forest model.

```{r fig-uncertainty, echo = FALSE, eval = TRUE, fig.cap="Conceptual illustration of measurement and prediction uncertainty. Narrow distributions represent wet-chemistry observations with lower analytical uncertainty, whereas wider distributions represent DRS predictions with larger associated standard deviations. Horizontal arrows indicate one standard deviation around the estimated value.", fig.width=6.5, fig.height=4.5}

library(ggplot2)

wet_mean <- 18.5
wet_sd <- 0.4
spec_mean <- 17.8
spec_sd <- 1.2

x <- seq(
  min(wet_mean - 4*wet_sd, spec_mean - 4*spec_sd),
  max(wet_mean + 4*wet_sd, spec_mean + 4*spec_sd),
  length.out = 1000
)

col_wet <- "#4E84A8" # steel blue
col_spec <- "#C1614A" # muted terracotta

df <- rbind(
  data.frame(x = x, density = dnorm(x, wet_mean, wet_sd), method = "Wet chemistry"),
  data.frame(x = x, density = dnorm(x, spec_mean, spec_sd), method = "DRS predictions")
)

means_df <- data.frame(
  mean = c(wet_mean, spec_mean),
  sd = c(wet_sd, spec_sd),
  method = c("Wet chemistry", "DRS predictions")
)

ymax <- max(df$density)

ggplot(df, aes(x = x, y = density, colour = method, fill = method)) +
  
  geom_rect(
    data = means_df,
    aes(xmin = mean - sd, xmax = mean + sd, ymin = 0, ymax = Inf, fill = method),
    alpha = 0.12, inherit.aes = FALSE
  ) +
  
  geom_area(alpha = 0.20, position = "identity") +
  geom_line(linewidth = 1.0) +
  
  geom_vline(
    data = means_df,
    aes(xintercept = mean, colour = method),
    linetype = "dashed", linewidth = 0.7, show.legend = FALSE
  ) +
  
  # sigma_DRS arrow — placed low, well left of the wet-chem peak
  annotate(
    "segment",
    x = spec_mean - spec_sd, xend = spec_mean + spec_sd,
    y = ymax * 0.38, yend = ymax * 0.38,
    colour = col_spec, linewidth = 0.8,
    arrow = arrow(ends = "both", length = unit(0.12, "cm"), type = "closed")
  ) +
  annotate(
    "text",
    x = spec_mean - spec_sd - 0.15, y = ymax * 0.38,
    label = expression(sigma[DRS]),
    colour = col_spec, size = 4.2, hjust = 1
  ) +
  
  # sigma_wet arrow — placed high, anchored right of its own peak
  annotate(
    "segment",
    x = wet_mean - wet_sd, xend = wet_mean + wet_sd,
    y = ymax * 0.72, yend = ymax * 0.72,
    colour = col_wet, linewidth = 0.8,
    arrow = arrow(ends = "both", length = unit(0.12, "cm"), type = "closed")
  ) +
  annotate(
    "text",
    x = wet_mean + wet_sd + 0.15, y = ymax * 0.72,
    label = expression(sigma["wet chem."]),
    colour = col_wet, size = 4.2, hjust = 0
  ) +
  
  scale_colour_manual(
    values = c("Wet chemistry" = col_wet, "DRS predictions" = col_spec)
  ) +
  scale_fill_manual(
    values = c("Wet chemistry" = col_wet, "DRS predictions" = col_spec)
  ) +
  
  labs(
    x = "Soil property value", 
    y = "Probability density", 
    colour = NULL, 
    fill = NULL
  ) +
  
  theme_classic(base_size = 13) +
  theme(
    legend.position = "top",
    legend.key.size = unit(0.9, "lines"),
    legend.text = element_text(size = 11),
    axis.line = element_line(colour = "grey40"),
    panel.grid.major.y = element_line(colour = "grey92", linewidth = 0.4), 
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.border = element_rect(colour = "grey40", fill = NA, linewidth = 0.5)
  )
```



## Part II – Practical application  

### Workflow overview  
In this practical section, we illustrate how DRS-derived soil property estimates can complement conventional laboratory analyses in the production of digital soil maps. The methodology is demonstrated using the Kansas, USA, dataset introduced in Chapter \@ref(digital-soil-mapping). In addition to the soil property variables used in the DSM example, this dataset includes mid-infrared DRS measurements that were not used previously.

Figure \@ref(fig:fig-dsm-workflow) presents the conceptual workflow used in this section. The colors distinguish the main groups of steps:

- <span style="color:#4E84A8; font-weight:700;">Blue boxes</span> represent field sampling and laboratory reference analysis. These steps provide the measured soil property data used as calibration targets, either to fit a DRS calibration model or to support DSM.

- <span style="color:#C1614A; font-weight:700;">Terracotta boxes</span> correspond to the DRS component. This includes spectral acquisition, calibration modeling, and the generation of spectroscopically predicted soil properties with associated uncertainty, expressed as the prediction error standard deviation. These predictions are treated as an additional source of soil information in the DSM step.

- <span style="color:#7B6D9E; font-weight:700;">Purple boxes</span> indicate the data integration stage. Here, laboratory observations and DRS predictions are combined into a single dataset.

- <span style="color:#4E8A6A; font-weight:700;">Green boxes</span> represent the spatial modeling stage, or DSM. Environmental covariates are used together with soil observations to fit spatial prediction models. The uncertainty associated with spectroscopic predictions is used to down-weight these observations relative to laboratory measurements. The final outputs are continuous soil property maps and corresponding uncertainty maps.

All main steps in this workflow are implemented and illustrated using R in the following subsections.

```{r fig-dsm-workflow, echo=FALSE, message=FALSE, warning=FALSE, out.width="80%", fig.cap="Conceptual workflow for integrating spectroscopically predicted soil data into digital soil mapping.", eval=TRUE}
g <- grViz("
digraph dsm_workflow {

  graph [layout = dot, rankdir = TB, nodesep = 0.3, ranksep = 0.2
         bgcolor = '#FFFFFF']

  node [shape = box,
        style = 'rounded,filled',
        penwidth = 1.5,
        fontname = 'Helvetica Neue',
        fontsize = 10,
        margin = 0.2]

  edge [arrowsize = 0.7,
        penwidth = 1.2,
        color = '#B0B8C1']

  sampling [label = 'Soil sampling',                              fillcolor = '#D6E8F2', color = '#4E84A8', fontcolor = '#1A3D52']
  lab      [label = 'Lab analysis',                               fillcolor = '#D6E8F2', color = '#4E84A8', fontcolor = '#1A3D52']
  spectra  [label = 'Spectral scanning',                          fillcolor = '#F2DDD9', color = '#C1614A', fontcolor = '#5C2518']
  calib    [label = 'Calibration model',                          fillcolor = '#F2DDD9', color = '#C1614A', fontcolor = '#5C2518']
  pred     [label = 'Spectroscopic predictions\\n(+ uncertainty)', fillcolor = '#F2DDD9', color = '#C1614A', fontcolor = '#5C2518']
  harmon   [label = 'Data harmonisation\\n/ weighting',           fillcolor = '#E2DDED', color = '#7B6D9E', fontcolor = '#312650']
  covars   [label = 'Environmental covariates',                   fillcolor = '#D8E8DF', color = '#4E8A6A', fontcolor = '#1A3D2A']
  dsm      [label = 'DSM model',                                  fillcolor = '#D8E8DF', color = '#4E8A6A', fontcolor = '#1A3D2A']
  maps     [label = 'Predicted soil maps',                        fillcolor = '#D8E8DF', color = '#4E8A6A', fontcolor = '#1A3D2A']

  sampling -> lab
  sampling -> spectra

  lab -> calib
  spectra -> calib

  calib -> pred
  lab -> harmon
  pred -> harmon

  covars -> dsm
  harmon -> dsm

  dsm -> maps
}
")
out_png <- "images/module4/fig_dsm_workflow.png"
tmp_svg <- tempfile(fileext = ".svg")
export_svg(g) |> writeLines(tmp_svg)
rsvg_png(tmp_svg, out_png, width = 1400)
knitr::include_graphics(out_png)
```

### Generating DRS-derived soil property observations for DSM

In this section, we convert DRS measurements into soil property estimates that can be used as additional observations in DSM. The key requirement is that, for each soil sample, the spectroscopic modeling step provides both a predicted value of the target soil property and an estimate of prediction uncertainty. This follows the approach introduced in the Soil Spectroscopy Training Material (@Wadoux2025FAO, Section 4.3.3), where uncertainty is quantified at the sample level.

For DSM applications, these outputs can be treated similarly to laboratory observations. However, unlike laboratory data, DRS predictions are associated with explicit model-based uncertainty. This uncertainty information can later be used during data integration and spatial modeling.

In the following subsections, we demonstrate how to (i) import and visualize the spectral data, (ii) apply basic spectral preprocessing, (iii) build a spectroscopic prediction model with uncertainty, (iv) define laboratory measurement uncertainty, and (v) harmonize soil property predictions and uncertainties to a standard depth interval.

#### Load and plot the DRS data

First, we import the DRS dataset and prepare it for analysis. This dataset contains absorbance measurements collected in the mid-infrared (MIR) region of the electromagnetic spectrum, over the wavenumber range starting from apprximately 4000 cm$^{-1}$ up to apprximately 400 cm$^{-1}$ at an approximated resolution of 1.93 cm$^{-1}$. This resulted in 1764 spectral variables. Note that this region corresponds to wavelengths of approximately 2500–25000 nm. These measurements were collected under laboratory conditions using air-dried soil samples. Additional details on the DRS measurements are provided in @dangal2019accurate.


> <span  style="font-weight:700;">NOTE:</span> _MIR spectra show strong, well-defined absorption features associated with molecular vibrations in soil minerals and organic compounds. This makes MIR spectroscopy sensitive to properties such as soil organic carbon, clay minerals, carbonates, and some nutrients.
Although this example uses MIR spectra, the workflow can also be applied to vis–NIR or SWIR data, provided suitable calibration models and uncertainty estimates are available._


Let's start by loading the data:

```{r message=FALSE, warning=FALSE, eval=TRUE}
library(readxl)
library(dplyr)

# Read MIR spectral dataset (Kansas example)
original_dat <- read_excel(
  "01_data/module1/training_data/data/MIR_KANSAS_data.xlsx",
  .name_repair = "minimal",
  na = c("", "NA", "N/A", "NaN"),
  progress = FALSE
)

original_dat <- as.data.frame(original_dat)

# dimensions of the dataset (number of samples and variables)
dim(original_dat)

# number of duplicate sample IDs (i.e. repeated measurements for 
# the same soil sample)a
sum(duplicated(original_dat$smp_id))
```
The loaded dataset contains both DRS variables and non-DRS variables. The first step is therefore to separate the spectral data from the rest of the soil dataset. To do this, we create a data frame containing the non-spectral variables and store the DRS variables as a spectral matrix. This structure allows the spectra to be handled separately for preprocessing and modeling, while keeping them linked to the corresponding soil samples.

The spectral variables can be identified from the column names. These can be inspected using `colnames(dat)`. In this dataset, spectral columns follow a consistent naming pattern: they start with the prefix `X`, followed by numeric values representing the original wavenumbers. For example, a column named `X3999.7` corresponds to a spectral variable measured at approximately 3999.7 cm$^{-1}$.

We use this pattern to separate the spectral variables from the remaining soil data. The `X` prefix is then removed so that the column names can be converted into numeric wavenumber values. These values are stored in `original_wavs` and used later for plotting, resampling, and spectral preprocessing.

```{r message=FALSE, warning=FALSE, eval=TRUE}
# Identify spectral columns
# Spectral variables start with "X" followed by at least three digits.
spc_cols <- grep("^X[0-9]{3,}", colnames(original_dat))

# Extract the spectral matrix
spc_original <- original_dat[, spc_cols]

# Remove the "X" prefix from spectral column names
colnames(spc_original) <- gsub("^X", "", colnames(spc_original))

# Convert spectral column names to numeric wavenumber values
original_wavs <- as.numeric(colnames(spc_original))

# Check the spectral range
range(original_wavs)
```

Because the spectra are recorded at high spectral resolution, the spectral matrix contains many highly correlated variables. Before modeling, we therefore resample the spectra to a coarser wavenumber grid. This reduces computational demand while preserving the main absorption features. In this case, we resample the spectra to a resolution of 8 cm$^{-1}$ using `prospectr::resample()`. Since the original spectra do not extend exactly to 4000 cm$^{-1}$, we avoid extrapolation (highly recommended for all cases) and resample only up to 3992 cm$^{-1}$. This results in a new spectral matrix with 425 spectral variables, which is more manageable for modeling while still capturing the key spectral information:

```{r message=FALSE, warning=FALSE, eval=TRUE}
library(prospectr)
# Define a new regular spectral grid for resampling
wavs <- seq(600, 3992, by = 8)
# Resample spectra to the new wavelength grid
spc_resampled <- prospectr::resample(
  spc_original,
  wav = original_wavs,
  new.wav = wavs
)
# check the dimensions
dim(spc_resampled)
```

Now we can plug this resampled spectral matrix into the non-DRS part of the dataframe:
```{r message=FALSE, warning=FALSE, eval=TRUE, results='hide'}
# extract non-spectral variables into a new dataframe
repeated_dat <- original_dat[, -spc_cols]

# in the main dataframe create a new variable called spc 
# containing the matrix of spectra 
repeated_dat$spc <- spc_resampled

# make some checks
colnames(repeated_dat)
dim(repeated_dat)
dim(repeated_dat$spc)
```

The data contains repeated measurements for each soil sample, which is common in spectroscopic datasets to account for measurement variability. To simplify the analysis, we start by averaging replicated spectral measurements so that each soil sample is represented by a single spectrum.

```{r message=FALSE, warning=FALSE, eval=TRUE, results = 'hide'}
# Average replicate spectra by sample ID
spc_avg <- aggregate(
  repeated_dat$spc,
  by = list(smp_id = repeated_dat$smp_id),
  FUN = mean
)
# order the resuslts by sample ID to ensure correct alignment with the main 
# dataset later on
spc_avg <- spc_avg[order(spc_avg$smp_id),  ]

# aggregate() prepends smp_id as first column; extract the spectral matrix
rownames(spc_avg) <- spc_avg$smp_id
spc_avg <- as.matrix(spc_avg[, -1])

# Drop duplicate non-spectral rows (keep first occurrence per sample)
dat <- repeated_dat[!duplicated(repeated_dat$smp_id), ]
dat <- dat[order(dat$smp_id), ]

# Verify that sample IDs are aligned before overwriting spectra
stopifnot(all(rownames(spc_avg) == dat$smp_id))

# Overwrite spectra with the averaged matrix
dat$spc <- spc_avg

# check dimensions of the final dataset
dim(dat)
```

Now let's plot the spectra (Figure \@ref(fig:fig-spectra)) that we have so far:
```{r fig-spectra, message=FALSE, warning=FALSE, eval=TRUE, out.width="80%", fig.cap="Resampled and aggreagted MIR sectra."}
matplot(
  wavs,
  t(dat$spc),
  type = "l",
  lty = 1,
  xlim = c(4000, 500),
  xlab = expression(Wavenumber~(cm^{-1})),
  ylab = "Absorbance",
  col = rgb(0.8, 0, 0, 0.3)
)
grid(lty = 1)
```

A unique profile identifier is also created to group horizons belonging to the same soil profile:

```{r message=FALSE, warning=FALSE, eval=TRUE, results = 'hide'}
# Create unique profile IDs for each coordinate pair
coords <- paste(dat$Long_Site.x, dat$Lat_Site.x, sep = "_")
prof_num <- match(coords, unique(coords))

dat$ProfID <- sprintf("PROF%04d", prof_num)
```

#### Preprocessing of spectra

Raw soil spectra often require preprocessing before modeling. The objective is to reduce instrumental or measurement artifacts while preserving the chemical information relevant for prediction.

A first step is always to inspect the spectra visually. This helps identify evident noise, baseline shifts, outliers, or unstable regions at the edges of the spectral range. Edge noise can occur because DRS detectors often have lower sensitivity near the limits of their operating range. When this occurs, it is common practice to trim the affected regions before modeling. In this example, we do not trim the spectra, but this should be considered for other datasets.

Spectra can also be affected by variation in instrument response, sample presentation, particle size, packing density, and measurement conditions. These effects may introduce additive or multiplicative changes in the spectra, even when the same sample is measured more than once. Common preprocessing methods used to reduce these effects include:

- **Smoothing**, such as moving average or Savitzky–Golay filtering, to reduce high-frequency noise.
- **Scatter correction**, such as standard normal variate (SNV) or multiplicative scatter correction (MSC), to reduce multiplicative effects.
- **Derivatives**, often combined with smoothing, to reduce baseline effects and enhance spectral features.

Derivatives should be used with caution because they can also amplify noise. The choice of preprocessing method should therefore depend on the spectral region, the quality of the measurements, and the modeling objective.

> <span  style="font-weight:700;">NOTE:</span> _Scatter correction methods may sometimes appear to reduce model accuracy during internal validation, especially when calibration and validation samples come from the same instrument, preparation protocol, and measurement conditions. In such cases, raw or lightly smoothed spectra may preserve dataset-specific variation that improves apparent performance. However, this variation may not be chemically meaningful and can make the model more sensitive to changes in particle size, packing density, sample presentation, instrument response, or acquisition conditions. Carefully selected scatter correction methods, such as SNV or MSC, may therefore produce models that are slightly less accurate in the short term but more robust when applied to new samples, instruments, or measurement settings. For this reason, scatter correction should be evaluated not only in terms of cross-validation accuracy, but also in terms of expected model transferability and long-term stability._

In this example, we first apply standard normal variate (SNV) correction to reduce multiplicative scatter effects, followed by a Savitzky–Golay first derivative to reduce baseline effects and enhance spectral features. In `savitzkyGolay()`, `m = 1` specifies the first derivative, `p = 2` sets the polynomial order used within the moving window, and `w = 11` defines the window size in spectral variables. Note that the optimal preprocessing strategy should still be evaluated for each dataset.

```{r message=FALSE, warning=FALSE}
# apply standard normal variate (SNV) correction followed by 
# a Savitzky–Golay first derivative
dat$spc_processed  <- dat$spc |> 
  prospectr::standardNormalVariate() |> 
  prospectr::savitzkyGolay(m = 1, p = 1, w = 11)

# get the wavenumber values for the processed spectra
wavs_processed <- as.numeric(colnames(dat$spc_processed))
```


Now let's visualise the processed spectra(Figure \@ref(fig:fig-spectra-processed)) that we have so far:
```{r fig-spectra-processed, message=FALSE, warning=FALSE, eval=TRUE, out.width="80%", fig.cap = "Processed MIR spectra after standard normal variate correction and Savitzky–Golay first-derivative transformation."}
matplot(
  wavs_processed,
  t(dat$spc_processed),
  type = "l",
  lty = 1,
  xlim = c(4000, 500),
  xlab = expression(Wavenumber~(cm^{-1})),
  ylab = "First derivative of SNV-corrected absorbance", 
  col = rgb(0.3, 0, 0.8, 0.3)
)
grid(lty = 1)
```


This basic exploratory step helps verify that the spectra have been imported and processed correctly. In a full analysis, this stage can also be used to screen for unusual spectra or potential measurement issues before fitting the calibration model. The dataset is now ready for statistical modeling.


#### Spectroscopic modelling with uncertainty

The next step is to build a model that predicts SOC from the preprocessed spectra. Here we use a random forest approach, called quantile regression forest. The same model type will then be used for the DSM. This model allows uncertainty to be estimated from the distribution of model predictions.

To obtain a realistic assessment of model performance, cross-validation is done at the soil profile level. This avoids using horizons from the same profile in both training and testing subsets. For each validation fold, the model is trained on a subset of profiles and used to predict SOC for the remaining profiles.

Prediction uncertainty is quantified by the model using the conditional distribution of the predictions. This has been extensively described for spectroscopic data in @wadoux2025uncertainty. From this model distribution, a mean predicted value, a prediction variance, and a prediction standard deviation are calculated for each soil sample. These quantities represent the required outputs of the spectroscopic modelling step for DSM.

Model performance can be first visualised by plotting predicted versus measured SOC values in a scatterplot. Validation statistics can be obtained from the function `ranger`` internal structure, where a R$^2$ is reported from a cross-validation. Error bars in the scatterplot representing the prediction standard deviation, they illustrate the level of confidence associated with each prediction.

```{r message=FALSE, warning=FALSE}
# -----------------------------------------------------------
# Quantile Random Forest with 10-fold cross-validation
# grouped by soil profile
# -----------------------------------------------------------

library(dplyr)
library(ranger)
library(matrixStats)

set.seed(123)

# -----------------------------------------------------------
# Assign entire soil profiles to CV folds
# This avoids splitting horizons from the same profile
# across training and testing data
# -----------------------------------------------------------

profile_ids <- unique(dat$ProfID)
n_profiles  <- length(profile_ids)

fold_table <- data.frame(
  ProfID = profile_ids,
  fold   = sample(rep(1:10, length.out = n_profiles))
)

# Add fold membership to the main dataset
dat <- dat %>%
  left_join(fold_table, by = "ProfID")

# -----------------------------------------------------------
# Prepare storage vectors for CV predictions and uncertainty
# -----------------------------------------------------------

SOC_predRF <- rep(NA_real_, nrow(dat))  # mean prediction
SOC_varRF  <- rep(NA_real_, nrow(dat))  # predictive variance
SOC_sdRF   <- rep(NA_real_, nrow(dat))  # predictive standard deviation

# -----------------------------------------------------------
# Run 10-fold cross-validation
# -----------------------------------------------------------

for (k in 1:10) {
  
  cat("\rProcessing fold", k, "of 10")
  
  # Test set: all samples belonging to fold k
  test_rows <- which(dat$fold == k)
  
  # Training set: all remaining samples
  train_rows <- which(dat$fold != k)
  
  # Fit quantile random forest on training data
  model <- ranger(
    x = dat$spc[train_rows, ],
    y = dat$SOC[train_rows],
    quantreg = TRUE,
    num.trees = 100,
    sample.fraction = 1,
    replace = TRUE,
    min.node.size = 2
  )
  
  # Predict by drawing samples from the conditional distribution
  pred_matrix <- predict(
    model,
    data = dat$spc[test_rows, ],
    type = "quantiles",
    what = function(x) sample(x, 100, replace = TRUE)
  )$predictions
  
  # Store prediction summary statistics
  SOC_predRF[test_rows] <- rowMeans(pred_matrix)
  SOC_varRF[test_rows]  <- rowVars(pred_matrix)
  SOC_sdRF[test_rows]   <- sqrt(SOC_varRF[test_rows])
}

# -----------------------------------------------------------
# Save predictions back into the dataset
# -----------------------------------------------------------

dat$SOC_predRF <- SOC_predRF
dat$SOC_varRF  <- SOC_varRF
dat$SOC_sdRF   <- SOC_sdRF
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# -----------------------------------------------------------
# Plot predicted vs measured SOC
# Horizontal error bars represent predictive uncertainty
# -----------------------------------------------------------

plot(
  x = dat$SOC_predRF,
  y = dat$SOC,
  xlab = "Predicted Organic Carbon (%)",
  ylab = "Measured Organic Carbon (%)",
  xlim = range(dat$SOC_predRF, dat$SOC, na.rm = TRUE),
  ylim = range(dat$SOC_predRF, dat$SOC, na.rm = TRUE),
  pch = 16,
  col = "white"
)

# Add horizontal error bars (± 1 predictive SD)
arrows(
  x0 = dat$SOC_predRF - dat$SOC_sdRF,
  y0 = dat$SOC,
  x1 = dat$SOC_predRF + dat$SOC_sdRF,
  y1 = dat$SOC,
  code = 3,
  angle = 90,
  length = 0.02,
  col = rgb(1, 0.65, 0, 0.7)
)

# Add prediction points
points(
  x = dat$SOC_predRF,
  y = dat$SOC,
  pch = 16,
  col = rgb(0, 0, 0, 0.7)
)

# Add 1:1 reference line
abline(0, 1, col = "red")

# Add grid
grid()
```


#### Laboratory measurement uncertainty

Soil organic carbon (SOC) concentrations were determined using high-temperature dry combustion. This method is widely regarded as one of the most precise laboratory approaches for SOC determination and is commonly used as a reference method.

Analytical precision for combustion-based SOC measurements is typically on the order of a few percent relative standard deviation, according to laboratory validation studies, generally around 2--3\%. Based on this information, it is possible to estimate the standard deviation associated with analytical SOC measurements.

Measurement uncertainty in SOC analysis is usually assumed to be proportional to the magnitude of the measurement. Accordingly, the laboratory measurement error can be expressed as $\sigma_{\text{laboratory}} = CV \cdot y_i$ where $CV$ is the coefficient of variation and $y_i$ is the measured SOC value.

```{r}
# -----------------------------------------------------------
# Approximate laboratory measurement uncertainty for SOC
# -----------------------------------------------------------
# We assume a relative analytical error of 3%.
# Therefore, the measurement error standard deviation is
# calculated as 3% of the measured SOC value.

dat$SOC_sdLAB <- dat$SOC * 0.03
```


#### Depth harmonisation of predictions and uncertainties

Digital soil mapping often requires soil properties to be expressed for standard depth intervals, such as 0–30 cm. However, field observations and spectroscopic predictions are typically available at horizon level with varying thickness. 

In this step, soil profiles are converted into a structured profile object that allows depth-based calculations. Predicted SOC values are then aggregated to the target depth interval using depth-weighted averaging, taking into account the proportion of each horizon that falls within the interval.

Uncertainty values are harmonised in a similar way. A depth-weighted approach is used to combine horizon-level prediction uncertainties and laboratory uncertainties into a single standard deviation for the 0–30 cm layer. Profiles that do not fully cover the target depth interval are excluded from the summary.

The final output of this preparation stage is therefore a dataset containing, for each soil profile location:
* a depth-harmonised predicted SOC value,
* a corresponding spectroscopic prediction uncertainty,
* an estimated laboratory measurement uncertainty, and
* spatial coordinates for mapping.

These prepared variables can now be used as inputs in the DSM modelling.

```{r}
# -----------------------------------------------------------
# Aggregate SOC predictions and uncertainties to 0-30 cm
# using profile information
# -----------------------------------------------------------

library(aqp)
library(dplyr)

# Remove spectral matrix columns before converting to an aqp object
# This keeps only tabular horizon-level information
dat <- dat[, !names(dat) %in% c("spc", "spc_raw")]

# Build a SoilProfileCollection object
# ProfID identifies each profile, and the top/bottom columns define horizons
depths(dat) <- ProfID ~ Top_depth_cm.x + Bottom_depth_cm.x

# -----------------------------------------------------------
# Depth-weighted mean SOC prediction for 0-30 cm
# -----------------------------------------------------------

soc_mean_0_30 <- slab(
  dat,
  fm = ~ SOC_predRF,
  slab.structure = c(0, 30)
)

# -----------------------------------------------------------
# Helper function:
# Compute a depth-weighted standard deviation over 0-30 cm
# Only the horizon portions overlapping the target interval are used
# Profiles that do not fully cover 0-30 cm return NA
# -----------------------------------------------------------

weighted_sd_0_30 <- function(top, bottom, sd, z1 = 0, z2 = 30) {
  w <- pmax(0, pmin(bottom, z2) - pmax(top, z1))
  keep <- w > 0 & !is.na(sd)
  
  w  <- w[keep]
  sd <- sd[keep]
  
  if (sum(w) < (z2 - z1)) return(NA_real_)
  
  sqrt(sum((w * sd)^2)) / sum(w)
}

# -----------------------------------------------------------
# Compute profile-level uncertainty for 0-30 cm
# for both RF prediction uncertainty and lab measurement uncertainty
# -----------------------------------------------------------

soc_sd_0_30 <- horizons(dat) %>%
  as.data.frame() %>%
  group_by(ProfID) %>%
  summarise(
    SOC_sdRF_0_30 = weighted_sd_0_30(
      Top_depth_cm.x,
      Bottom_depth_cm.x,
      SOC_sdRF,
      0, 30
    ),
    SOC_sdLAB_0_30 = weighted_sd_0_30(
      Top_depth_cm.x,
      Bottom_depth_cm.x,
      SOC_sdLAB,
      0, 30
    ),
    .groups = "drop"
  )

# -----------------------------------------------------------
# Compute profile-level mean predicted SOC for 0-30 cm
# using overlap with the target interval as weights
# Profiles that do not fully cover 0-30 cm return NA
# -----------------------------------------------------------

soc_mean_vals <- profileApply(dat, function(p) {
  x <- horizons(p)
  
  overlap <- pmax(0, pmin(x$Bottom_depth_cm.x, 30) - pmax(x$Top_depth_cm.x, 0))
  keep <- overlap > 0 & !is.na(x$SOC_predRF)
  
  if (sum(overlap[keep]) < 30) return(NA_real_)
  
  sum(overlap[keep] * x$SOC_predRF[keep]) / sum(overlap[keep])
})

# -----------------------------------------------------------
# Combine mean prediction and uncertainty into one table
# -----------------------------------------------------------

soc_0_30 <- soc_sd_0_30 %>%
  mutate(SOC_predRF_0_30 = soc_mean_vals) %>%
  select(ProfID, SOC_predRF_0_30, SOC_sdRF_0_30, SOC_sdLAB_0_30)

# -----------------------------------------------------------
# Extract one set of coordinates per profile
# -----------------------------------------------------------

dat_sub <- as.data.frame(dat)[, c("ProfID", "SOC", "Lat_Site.x", "Long_Site.x")]
coords  <- dat_sub[!duplicated(dat_sub$ProfID), ]

# -----------------------------------------------------------
# Attach coordinates to the 0-30 cm summary table
# -----------------------------------------------------------

soc_0_30_xy <- merge(
  soc_0_30,
  coords,
  by = "ProfID",
  all.x = TRUE
)

# Display the first profile-level results
print(head(soc_0_30_xy))
```


### Preparation of the covariates

At this stage, the environmental covariates are prepared exactly as in the training material, so the same workflow can be followed here. The objective is to align the soil observations with the raster covariate stack and extract, for each sampling location, the values of all environmental predictors that will later be used in the modelling step.

The procedure involves four main operations:
* loading the raster stack of environmental covariates,
* converting the soil profile summaries into a spatial point object,
* reprojecting the points so that they match the coordinate reference system of the covariates,
* extracting raster values at each soil observation location and attaching them to the point data.

This produces a spatial dataset in which each soil observation is associated with the full set of environmental predictors.
```{r message=FALSE, warning=FALSE}
library(terra)

# -----------------------------------------------------------
# Load environmental covariates
# -----------------------------------------------------------

covs <- rast(
  "01_data/module1/training_data/data/Environmental_Covariates_250m_KANSAS.tif"
)
cov_names <- names(covs)

# -----------------------------------------------------------
# Convert the 0-30 cm soil summary table to spatial points
# -----------------------------------------------------------

soil_pts <- vect(
  soc_0_30_xy,
  geom = c("Long_Site.x", "Lat_Site.x"),
  crs = "EPSG:4326"
)

# Reproject points to match the raster covariate stack
soil_pts <- project(soil_pts, covs)

# -----------------------------------------------------------
# Extract covariate values at each soil point
# -----------------------------------------------------------

covs_extracted <- extract(covs, soil_pts)

# Remove the ID column returned by extract() and attach predictors
soil_pts <- cbind(soil_pts, covs_extracted[, -1])

# Optional check: plot sample locations
plot(soil_pts, cex = 0.7, col = "red")

```

### DSM model training with spectroscopic data

In this section we demonstrate how spectroscopic predictions can be integrated into the DSM modelling. The idea is that each observation, whether originating from laboratory analysis or from spectroscopy, is associated with a sample-specific uncertainty estimate.
This uncertainty can be used to control the relative influence of each observation during model training.

As introduced in Part I (Section 3), a practical way to achieve this is through case weighting. Observations with lower uncertainty are given higher statistical weight, while observations with higher uncertainty contribute less to the fitted model. This approach allows spectroscopic predictions to supplement laboratory data without treating both sources as equally reliable. 

In the following example, SOC is mapped using quantile regression forests, following a workflow similar to previous DSM examples. The steps include deriving uncertainty-based weights, fitting models with and without spectroscopic data, and comparing the resulting spatial predictions.

#### Deriving observation weights

Weights are derived from the uncertainty estimates obtained in the previous section. A simple and commonly used formulation is to define weights as the inverse of the variance associated with each observation.

This means that observations with smaller prediction or measurement variance receive larger weights. Conversely, observations with large uncertainty contribute less to model calibration.

Separate raw weights are therefore computed for:
* spectroscopic predictions (based on prediction standard deviation), and
* laboratory measurements (based on analytical uncertainty).

Inspecting the distribution of these weights helps to understand the relative precision of the two data sources and to detect potential extreme values before model fitting.

```{r message=FALSE, warning=FALSE}
# -----------------------------------------------------------
# Define inverse-variance weights for modelling
# -----------------------------------------------------------

soil_pts$w_RF_raw  <- 1 / (soil_pts$SOC_sdRF_0_30^2)
soil_pts$w_LAB_raw <- 1 / (soil_pts$SOC_sdLAB_0_30^2)

# Optional: check weight distributions
summary(soil_pts$w_RF_raw)
summary(soil_pts$w_LAB_raw)
```

#### Baseline DSM model using laboratory data

Before integrating spectroscopic information, a baseline DSM model can be fitted using laboratory SOC observations only. Environmental covariates representing soil-forming factors are used as predictors in a random forest model.

This step provides a reference spatial prediction of SOC across the study area. The resulting map can later be compared with models that include spectroscopic information. Such comparisons help to evaluate whether the additional data source improves spatial patterns or prediction stability.

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(ranger)
library(terra)

# -----------------------------------------------------------
# Prepare modelling table
# Keep coordinates, response, uncertainty columns, and predictors
# -----------------------------------------------------------

soil_df <- as.data.frame(soil_pts, geom = "XY")[,
  c("x", "y", "SOC", "SOC_predRF_0_30", "w_RF_raw", "w_LAB_raw", cov_names)
]

# -----------------------------------------------------------
# Fit baseline random forest model using laboratory SOC
# -----------------------------------------------------------

form <- as.formula(
  paste("SOC ~", paste(cov_names, collapse = " + "))
)

mod <- ranger(
  formula = form,
  data = soil_df,
  num.trees = 100
)

# -----------------------------------------------------------
# Predict SOC across the covariate raster stack
# -----------------------------------------------------------

soc_map <- predict(covs, mod, na.rm = TRUE)

# -----------------------------------------------------------
# Plot baseline SOC map
# -----------------------------------------------------------

plot(soc_map, main = "Baseline SOC map (lab observations only)")
```

#### DSM model training integrating spectroscopic predictions

To demonstrate the effect of spectroscopic supplementation, a mixed dataset is created. A subset of observations is replaced by spectroscopically predicted SOC values, together with their associated uncertainty-based weights.

An indicator variable can be retained to distinguish between laboratory and spectroscopic sources if needed for later analyses.

After removing incomplete or invalid cases, a weighted random forest model is fitted. The case weights ensure that observations contribute according to their estimated reliability, rather than simply according to their number.

Spatial predictions are then generated and compared with the baseline laboratory-only model. Mapping the difference between predictions helps to visualise how the inclusion of spectroscopic data influences the spatial distribution of SOC.

This exercise illustrates a practical strategy for integrating heterogeneous soil information sources in DSM. The same principles can be applied to other soil properties and modelling approaches, such as with geostatistics or other machine learning techniques commonly used in DSM.

```{r message=FALSE, warning=FALSE}
library(ranger)
library(terra)

set.seed(123)

# -----------------------------------------------------------
# Create a mixed dataset with 25% spectroscopic observations
# -----------------------------------------------------------

n <- nrow(soil_df)

# Randomly select 25% of observations to be replaced by RF predictions
idx_rf <- sample(seq_len(n), size = round(0.25 * n))

# Mixed response variable:
# keep lab SOC by default, replace selected points with RF-based SOC
soil_df$SOC_mix <- soil_df$SOC
soil_df$SOC_mix[idx_rf] <- soil_df$SOC_predRF_0_30[idx_rf]

# Matching weights:
# lab points keep lab weights, RF points receive RF-based weights
soil_df$w_mix <- soil_df$w_LAB_raw
soil_df$w_mix[idx_rf] <- soil_df$w_RF_raw[idx_rf]

# Optional indicator showing the source of each observation
soil_df$source <- "LAB"
soil_df$source[idx_rf] <- "RF"

# -----------------------------------------------------------
# Keep only valid rows for modelling
# -----------------------------------------------------------

vars_needed <- c("SOC_mix", "w_mix", cov_names)

soil_df2 <- soil_df[
  complete.cases(soil_df[, vars_needed]) &
    is.finite(soil_df$w_mix) &
    soil_df$w_mix > 0,
]

# -----------------------------------------------------------
# Fit weighted random forest model
# -----------------------------------------------------------

form_mix <- as.formula(
  paste("SOC_mix ~", paste(cov_names, collapse = " + "))
)

mod_mix25 <- ranger(
  formula = form_mix,
  data = soil_df2,
  case.weights = soil_df2$w_mix,
  num.trees = 100
)

# -----------------------------------------------------------
# Predict SOC map for the mixed scenario
# -----------------------------------------------------------

soc_map_mix25 <- predict(covs, mod_mix25, na.rm = TRUE)

# -----------------------------------------------------------
# Compare baseline and mixed maps
# -----------------------------------------------------------

par(mfrow = c(1, 3))

plot(soc_map, main = "SOC map: lab only")
plot(soc_map_mix25, main = "SOC map: 25% spectro")
plot(soc_map - soc_map_mix25, main = "Difference")
```

### Model validation

In this section we evaluate the predictive performance of the DSM models. Because the dataset combines laboratory measurements and spectroscopically derived estimates, model validation must explicitly account for observation-specific uncertainty. Ignoring these differences would treat all observations as equally reliable, which is inconsistent with the modelling strategy adopted in previous sections.

To address this, validation statistics are computed using uncertainty-based weights. These weights correspond to the inverse variance of each observation and therefore reflect the relative confidence in laboratory and spectroscopic data. Weighted validation provides a more realistic assessment of model performance when heterogeneous data sources are used.

#### Validation statistics

Three weighted performance metrics are considered. These metrics extend commonly used DSM validation statistics by incorporating observation weights.

The weighted mean error ($ME_w$) measures the average signed prediction bias, taking into account the reliability of each observation.
The weighted ME:
$$
ME_w =
\frac{\sum_{i=1}^{n} w_i \left( \hat{z}(s_i) - z_m(s_i) \right)}
{\sum_{i=1}^{n} w_i},
$$

The weighted root mean square error ($RMSE_w$) quantifies the overall magnitude of prediction errors while giving more importance to precise observations.
The weighted RMSE:
$$
RMSE_w =
\sqrt{
\frac{
\sum_{i=1}^{n} w_i \left( \hat{z}(s_i) - z_m(s_i) \right)^2
}{
\sum_{i=1}^{n} w_i
}
}
$$

Finally, the weighted Nash–Sutcliffe efficiency ($NSE_w$) evaluates predictive skill relative to the weighted mean of observed values.
Values of $NSE_w$ close to 1 indicate good predictive performance, whereas negative values indicate poor model skill.
The weighted NSE: 
$$
NSE_w =
1 -
\frac{
\sum_{i=1}^{n} w_i \left( \hat{z}(s_i) - z_m(s_i) \right)^2
}{
\sum_{i=1}^{n} w_i \left( z_m(s_i) - \bar{z}_m \right)^2
}
$$
Simple functions can be defined to compute these metrics from vectors of observations, predictions, and associated weights. These functions will be used throughout the validation exercises.




```{r}
# -----------------------------------------------------------
# Weighted validation metrics
# -----------------------------------------------------------

# Weighted Mean Error (bias)
wME <- function(obs, pred, w) {
  sum(w * (pred - obs), na.rm = TRUE) /
    sum(w, na.rm = TRUE)
}

# Weighted Root Mean Squared Error
wRMSE <- function(obs, pred, w) {
  sqrt(
    sum(w * (pred - obs)^2, na.rm = TRUE) /
      sum(w, na.rm = TRUE)
  )
}

# Weighted Nash–Sutcliffe Efficiency
wNSE <- function(obs, pred, w) {
  
  # weighted mean of observations
  zbar_w <- sum(w * obs, na.rm = TRUE) /
            sum(w, na.rm = TRUE)
  
  # weighted residual sum of squares
  num <- sum(w * (pred - obs)^2, na.rm = TRUE)
  
  # weighted total sum of squares
  den <- sum(w * (obs - zbar_w)^2, na.rm = TRUE)
  
  1 - num / den
}
```

#### Cross-validating the models

To obtain robust validation statistics, model performance is assessed using $k$-fold cross-validation. In this example, a mixed dataset is first constructed by combining laboratory SOC observations with a subset of spectroscopically predicted values. Matching uncertainty-based weights are assigned to each observation according to its data source.

The dataset is then partitioned into several folds. For each fold, the model is calibrated on the remaining data and used to predict SOC for the withheld observations. This procedure ensures that validation statistics reflect the model's ability to generalise to unseen locations.

Predicted values, observed values, and associated weights are stored and subsequently used to compute weighted ME, RMSE, and NSE. These statistics provide a quantitative summary of model performance when spectroscopic data are integrated into the calibration process.

```{r message=FALSE, warning=FALSE}
library(ranger)

set.seed(123)

# -----------------------------------------------------------
# Create mixed response (25% spectroscopic observations)
# -----------------------------------------------------------

n <- nrow(soil_df)

idx_rf <- sample(seq_len(n), size = round(0.25 * n))

soil_df$SOC_mix <- soil_df$SOC
soil_df$SOC_mix[idx_rf] <- soil_df$SOC_predRF_0_30[idx_rf]

soil_df$w_mix <- soil_df$w_LAB_raw
soil_df$w_mix[idx_rf] <- soil_df$w_RF_raw[idx_rf]

soil_df$source <- "LAB"
soil_df$source[idx_rf] <- "RF"

# -----------------------------------------------------------
# Keep valid rows for modelling
# -----------------------------------------------------------

vars_needed <- c("SOC_mix", "w_mix", cov_names)

soil_df2 <- soil_df[
  complete.cases(soil_df[, vars_needed]) &
    is.finite(soil_df$w_mix) &
    soil_df$w_mix > 0,
]

# -----------------------------------------------------------
# Define 10-fold cross-validation structure
# -----------------------------------------------------------

k <- 10
fold_id <- sample(rep(1:k, length.out = nrow(soil_df2)))

cv_pred <- rep(NA_real_, nrow(soil_df2))

form <- as.formula(
  paste("SOC_mix ~", paste(cov_names, collapse = " + "))
)

# -----------------------------------------------------------
# Run cross-validation loop
# -----------------------------------------------------------

for (fold in 1:k) {
  
  train_data <- soil_df2[fold_id != fold, ]
  test_data  <- soil_df2[fold_id == fold, ]
  
  mod_cv <- ranger(
    formula = form,
    data = train_data,
    case.weights = train_data$w_mix,
    num.trees = 100
  )
  
  cv_pred[fold_id == fold] <- predict(mod_cv, data = test_data)$predictions
}

# -----------------------------------------------------------
# Store CV predictions and compute weighted metrics
# -----------------------------------------------------------

cv_results <- data.frame(
  obs    = soil_df2$SOC_mix,
  pred   = cv_pred,
  w      = soil_df2$w_mix,
  source = soil_df2$source
)

ME_w   <- wME(cv_results$obs, cv_results$pred, cv_results$w)
RMSE_w <- wRMSE(cv_results$obs, cv_results$pred, cv_results$w)
NSE_w  <- wNSE(cv_results$obs, cv_results$pred, cv_results$w)

data.frame(
  Metric = c("wME", "wRMSE", "wNSE"),
  Value  = round(c(ME_w, RMSE_w, NSE_w), 3)
)

```

#### Comparing maps produced with different proportions of spectroscopic data

After model validation, spatial prediction maps can be generated. Maps derived from laboratory-only models can be compared with maps obtained from models that incorporate spectroscopic predictions. This visual comparison helps to identify changes in spatial patterns, prediction smoothness, and local variability.

Differences between maps may reflect both the increased data density and the influence of uncertainty-based weighting. Such comparisons are an important step in understanding the practical impact of spectroscopic data supplementation in DSM applications.

To visualise the effect of increasing the proportion of spectroscopic observations, several mixed-data scenarios can be fitted and mapped. In the example below, four proportions are considered: 10\%, 25\%, 50\%, and 75\% spectroscopic data. For each scenario, a weighted random forest is fitted, a SOC map is predicted, and the result is compared with the baseline map derived from laboratory observations only.

The final figure is organised in two rows:
* the top row shows the SOC maps for the four mixed scenarios, with sampling points coloured by source,
* the bottom row shows the difference between each mixed map and the baseline laboratory-only map.

```{r fig-mixed-maps-build, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE, fig.cap="SOC maps produced with increasing proportions of spectroscopic observations (top row) and their differences from the baseline laboratory-only SOC map (bottom row)."}
# -----------------------------------------------------------
# Sensitivity analysis: mapping SOC using different proportions
# of spectroscopic observations
# -----------------------------------------------------------

library(terra)
library(ranger)
library(ggplot2)
library(ggpubr)

set.seed(123)

# Proportions of spectroscopic observations to evaluate
props <- c(0.10, 0.25, 0.50, 0.75)

# Storage objects for maps and differences
map_plots  <- vector("list", length(props))
diff_plots <- vector("list", length(props))

# Temporary storage to harmonise colour scales across panels
map_vals   <- vector("list", length(props))
diff_vals  <- vector("list", length(props))

# -----------------------------------------------------------
# Loop over spectroscopic data proportions
# -----------------------------------------------------------

for (i in seq_along(props)) {
  
  p <- props[i]
  n <- nrow(soil_df)
  
  # Randomly select observations replaced by spectroscopic SOC
  idx_rf <- sample(seq_len(n), size = round(p * n))
  
  # Create mixed SOC response variable
  soil_df$SOC_mix <- soil_df$SOC
  soil_df$SOC_mix[idx_rf] <- soil_df$SOC_predRF_0_30[idx_rf]
  
  # Assign uncertainty-based weights
  soil_df$w_mix <- soil_df$w_LAB_raw
  soil_df$w_mix[idx_rf] <- soil_df$w_RF_raw[idx_rf]
  
  # Track data source for plotting
  soil_df$source <- "LAB"
  soil_df$source[idx_rf] <- "RF"
  
  # Keep only valid rows for modelling
  vars_needed <- c("SOC_mix", "w_mix", cov_names)
  
  soil_df2 <- soil_df[
    complete.cases(soil_df[, vars_needed]) &
      is.finite(soil_df$w_mix) &
      soil_df$w_mix > 0,
  ]
  
  # -----------------------------------------------------------
  # Fit weighted random forest model
  # -----------------------------------------------------------
  
  mod_mix <- ranger(
    formula = as.formula(
      paste("SOC_mix ~", paste(cov_names, collapse = " + "))
    ),
    data = soil_df2,
    case.weights = soil_df2$w_mix,
    num.trees = 50
  )
  
  # Predict spatial SOC map
  soc_map_mix <- predict(covs, mod_mix, na.rm = TRUE)
  
  # Compute difference from baseline laboratory-only map
  diff_map <- soc_map_mix - soc_map
  
  # Convert rasters to data frames for ggplot visualisation
  df_map  <- as.data.frame(soc_map_mix, xy = TRUE, na.rm = TRUE)
  df_diff <- as.data.frame(diff_map,    xy = TRUE, na.rm = TRUE)
  
  names(df_map)[3]  <- "value"
  names(df_diff)[3] <- "value"
  
  # Store values to define common colour scales later
  map_vals[[i]]  <- df_map$value
  diff_vals[[i]] <- df_diff$value
  
  # -----------------------------------------------------------
  # Build map panels
  # -----------------------------------------------------------
  
  map_plots[[i]] <- ggplot() +
    geom_raster(data = df_map, aes(x = x, y = y, fill = value)) +
    geom_point(
      data = soil_df2,
      aes(x = x, y = y, color = source),
      size = 0.6
    ) +
    scale_color_manual(values = c(LAB = "red", RF = "blue")) +
    coord_equal() +
    labs(
      title = paste0(round(100 * p), "% spectro"),
      fill  = "SOC",
      color = NULL
    ) +
    theme_minimal() +
    theme(
      axis.title = element_blank(),
      axis.text  = element_blank(),
      axis.ticks = element_blank(),
      panel.grid = element_blank()
    )
  
  diff_plots[[i]] <- ggplot() +
    geom_raster(data = df_diff, aes(x = x, y = y, fill = value)) +
    coord_equal() +
    labs(
      title = paste0("Difference ", round(100 * p), "%"),
      fill  = "Diff"
    ) +
    theme_minimal() +
    theme(
      axis.title = element_blank(),
      axis.text  = element_blank(),
      axis.ticks = element_blank(),
      panel.grid = element_blank()
    )
}

# -----------------------------------------------------------
# Define common colour scales across all panels
# -----------------------------------------------------------

map_range  <- range(unlist(map_vals),  na.rm = TRUE)
diff_range <- range(unlist(diff_vals), na.rm = TRUE)
diff_lim   <- max(abs(diff_range))

for (i in seq_along(props)) {
  
  map_plots[[i]] <- map_plots[[i]] +
    scale_fill_viridis_c(limits = map_range)
  
  diff_plots[[i]] <- diff_plots[[i]] +
    scale_fill_gradient2(
      limits   = c(-diff_lim, diff_lim),
      midpoint = 0
    )
}

# -----------------------------------------------------------
# Arrange final figure layout
# -----------------------------------------------------------

row1 <- ggarrange(
  plotlist = map_plots,
  ncol = 4, nrow = 1,
  common.legend = TRUE,
  legend = "top"
)

row2 <- ggarrange(
  plotlist = diff_plots,
  ncol = 4, nrow = 1,
  common.legend = TRUE,
  legend = "bottom"
)

final_fig <- ggarrange(row1, row2, ncol = 1, nrow = 2)
final_fig
```


#### Testing different ratios of spectroscopic and laboratory data

The influence of spectroscopic information on DSM can be further explored by systematically varying the proportion of spectroscopically inferred observations used during model calibration. Such a sensitivity analysis provides insight into how increasing reliance on spectroscopy may affect both the spatial pattern of predictions and the overall predictive performance of the model.

In this exercise, several scenarios are evaluated, ranging from situations where only a small fraction of the calibration dataset is derived from spectroscopy to cases where spectroscopic observations constitute a substantial component of the available information.
For each scenario, a weighted DSM model is fitted using uncertainty-based case weights, and spatial prediction maps are generated.

Difference maps relative to the laboratory-only baseline allow visual inspection of how predicted SOC distributions evolve as the balance between laboratory and spectroscopic data changes.

```{r,fig-sensitivity-metrics, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE, fig.cap="Sensitivity of weighted validation metrics to the proportion of spectroscopic observations. Points show individual repetitions, lines show the mean across repetitions, and ribbons represent one standard deviation."}
# -----------------------------------------------------------
# Sensitivity analysis: effect of increasing the proportion
# of spectroscopic observations on weighted model performance
# -----------------------------------------------------------

library(ranger)
library(ggplot2)
library(ggpubr)

set.seed(123)

# Number of observations and cross-validation settings
n <- nrow(soil_df)
k <- 10
n_rep <- 25

# Proportions of spectroscopic observations to test
perc_vec <- seq(1, 99, by = 5)

# Model formula
form <- as.formula(
  paste("SOC_mix ~", paste(cov_names, collapse = " + "))
)

# Storage table for all repetitions and proportions
results <- data.frame()

# -----------------------------------------------------------
# Loop over spectroscopic data proportions
# -----------------------------------------------------------

for (perc in perc_vec) {
  
  # Repeat each proportion several times
  # to account for randomness in the selected spectroscopic points
  for (rep in 1:n_rep) {
    
    # Randomly select the proportion of observations replaced
    # by spectroscopic SOC predictions
    idx_rf <- sample(seq_len(n), size = round((perc / 100) * n))
    
    # Create mixed SOC response
    soil_df$SOC_mix <- soil_df$SOC
    soil_df$SOC_mix[idx_rf] <- soil_df$SOC_predRF_0_30[idx_rf]
    
    # Assign the corresponding uncertainty-based weights
    soil_df$w_mix <- soil_df$w_LAB_raw
    soil_df$w_mix[idx_rf] <- soil_df$w_RF_raw[idx_rf]
    
    # Optional source indicator
    soil_df$source <- "LAB"
    soil_df$source[idx_rf] <- "RF"
    
    # Keep only complete and valid observations
    vars_needed <- c("SOC_mix", "w_mix", cov_names)
    
    soil_df2 <- soil_df[
      complete.cases(soil_df[, vars_needed]) &
        is.finite(soil_df$w_mix) &
        soil_df$w_mix > 0,
    ]
    
    # -----------------------------------------------------------
    # Create 10-fold cross-validation splits
    # -----------------------------------------------------------
    
    fold_id <- sample(rep(1:k, length.out = nrow(soil_df2)))
    cv_pred <- rep(NA_real_, nrow(soil_df2))
    
    # -----------------------------------------------------------
    # Run 10-fold cross-validation
    # -----------------------------------------------------------
    
    for (fold in 1:k) {
      
      train_data <- soil_df2[fold_id != fold, ]
      test_data  <- soil_df2[fold_id == fold, ]
      
      mod_cv <- ranger(
        formula = form,
        data = train_data,
        case.weights = train_data$w_mix,
        num.trees = 50,
        importance = "none",
        verbose = FALSE
      )
      
      cv_pred[fold_id == fold] <- predict(mod_cv, data = test_data)$predictions
    }
    
    # -----------------------------------------------------------
    # Compute weighted validation metrics for this repetition
    # -----------------------------------------------------------
    
    cv_results <- data.frame(
      obs  = soil_df2$SOC_mix,
      pred = cv_pred,
      w    = soil_df2$w_mix
    )
    
    results <- rbind(
      results,
      data.frame(
        perc_spectro = perc,
        repetition   = rep,
        wME   = wME(cv_results$obs, cv_results$pred, cv_results$w),
        wRMSE = wRMSE(cv_results$obs, cv_results$pred, cv_results$w),
        wNSE  = wNSE(cv_results$obs, cv_results$pred, cv_results$w)
      )
    )
  }
}

# -----------------------------------------------------------
# Summarise results across repetitions
# -----------------------------------------------------------

sum_mean <- aggregate(
  cbind(wME, wRMSE, wNSE) ~ perc_spectro,
  data = results,
  FUN = mean
)

sum_sd <- aggregate(
  cbind(wME, wRMSE, wNSE) ~ perc_spectro,
  data = results,
  FUN = sd
)

results_plot <- merge(
  sum_mean,
  sum_sd,
  by = "perc_spectro",
  suffixes = c("_mean", "_sd")
)

# -----------------------------------------------------------
# Plot weighted performance metrics
# Points show individual repetitions
# Line shows the mean across repetitions
# Ribbon shows ± 1 standard deviation
# -----------------------------------------------------------

p1 <- ggplot() +
  geom_point(
    data = results,
    aes(x = perc_spectro, y = wME),
    alpha = 0.5,
    size = 1
  ) +
  geom_ribbon(
    data = results_plot,
    aes(
      x = perc_spectro,
      ymin = wME_mean - wME_sd,
      ymax = wME_mean + wME_sd
    ),
    alpha = 0.2
  ) +
  geom_line(
    data = results_plot,
    aes(x = perc_spectro, y = wME_mean),
    linewidth = 0.8
  ) +
  labs(
    x = "% spectro data",
    y = "wME"
  ) +
  theme_minimal()

p2 <- ggplot() +
  geom_point(
    data = results,
    aes(x = perc_spectro, y = wRMSE),
    alpha = 0.5,
    size = 1
  ) +
  geom_ribbon(
    data = results_plot,
    aes(
      x = perc_spectro,
      ymin = wRMSE_mean - wRMSE_sd,
      ymax = wRMSE_mean + wRMSE_sd
    ),
    alpha = 0.2
  ) +
  geom_line(
    data = results_plot,
    aes(x = perc_spectro, y = wRMSE_mean),
    linewidth = 0.8
  ) +
  labs(
    x = "% spectro data",
    y = "wRMSE"
  ) +
  theme_minimal()

p3 <- ggplot() +
  geom_point(
    data = results,
    aes(x = perc_spectro, y = wNSE),
    alpha = 0.5,
    size = 1
  ) +
  geom_ribbon(
    data = results_plot,
    aes(
      x = perc_spectro,
      ymin = wNSE_mean - wNSE_sd,
      ymax = wNSE_mean + wNSE_sd
    ),
    alpha = 0.2
  ) +
  geom_line(
    data = results_plot,
    aes(x = perc_spectro, y = wNSE_mean),
    linewidth = 0.8
  ) +
  labs(
    x = "% spectro data",
    y = "wNSE"
  ) +
  theme_minimal()

# Final combined figure
perf_fig <- ggarrange(p1, p2, p3, ncol = 1, nrow = 3)
perf_fig

```


## Summary 

In this practical exercise we demonstrated how soil spectroscopic information can be integrated into a digital soil mapping workflow. Mid-infrared spectral data were first preprocessed and used to build a calibration model that provided, for each soil sample, both a predicted SOC value and a sample-specific estimate of prediction uncertainty. These spectroscopic predictions were then harmonised to a standard depth interval and combined with laboratory observations. Laboratory analytical uncertainty was also approximated so that both data sources could be treated consistently.

A key step of the workflow was the use of uncertainty-based case weights during spatial model training. By assigning weights proportional to the inverse of observation variance, laboratory data and spectroscopic predictions contributed to model calibration according to their relative reliability. Model validation was also performed using weighted performance metrics, ensuring that predictive skill was evaluated in a way that reflects heterogeneous data quality. Spatial prediction maps illustrated how supplementing laboratory data with spectroscopy can modify spatial patterns and potentially improve prediction stability, depending on the proportion and uncertainty of the additional data.

Although the example focused on SOC mapping with quantile regression forests, the same principles can be applied to other soil properties and modelling approaches. Only minor adaptations are required, such as changing the calibration model, the spatial prediction algorithm, or the method used to propagate uncertainty. More broadly, the workflow is a practical framework for integrating emerging proximal sensing data sources into DSM while maintaining a transparent treatment of data quality and prediction uncertainty.
