# Soil Spectroscopy for Digital Soil Mapping

*Alexandre Wadoux and Leonardo Ramirez-Lopez*

```{r setup, include=FALSE}
required_packages <- c(
  "terra", "ranger", "dplyr", "ggplot2",
  "prospectr", "matrixStats", "DiagrammeR",
  "DiagrammeRsvg", "rsvg"
)

dir.create("images/module4", recursive = TRUE, showWarnings = FALSE)

missing <- required_packages[!sapply(required_packages, requireNamespace, quietly = TRUE)]
if (length(missing) > 0) install.packages(missing)

library(terra)
library(ranger)
library(dplyr)
library(ggplot2)
library(prospectr)
library(matrixStats)
library(DiagrammeR)
library(DiagrammeRsvg)
library(rsvg)

knitr::opts_chunk$set(
  echo = TRUE,
  eval = FALSE,
  message = FALSE,
  warning = FALSE, 
  fig.align = 'center'
)
```

## Introduction 

Digital soil mapping (DSM) relies heavily on laboratory analyses to measure and characterize soil properties at sampled locations, which are then used as reference observations for training and validating spatial prediction models. Although highly accurate, conventional laboratory methods are slow and costly, creating a bottleneck that restricts timely data production for DSM. For example, conventional soil organic carbon reference methods, such as dry combustion and wet oxidation, require specialized infrastructure, trained personnel, and consumables, leading to long turnaround times and substantial per-sample costs [@wu2025assessing]. In this context, sensing technologies such as diffuse reflectance spectroscopy can complement laboratory analysis by reducing analytical costs and enabling soil property observations to be generated more rapidly and at greater spatial density. For these reasons, diffuse reflectance spectroscopy has gained substantial attention in the soil science community as a tool for enhancing soil characterization [@peng2025spectroscopic].

In soil applications, diffuse reflectance spectroscopy (DRS) is a maturing spectroscopic approach focused on extracting information from the interaction between soil samples and electromagnetic radiation. The chemical composition and physical structure of soil determine how electromagnetic radiation is absorbed and scattered. Consequently, soil composition and physical properties can be inferred by analyzing reflected energy. DRS sensors record reflected energy across multiple wavelengths, producing spectra that are typically converted from reflectance to absorbance. Soil property quantification with DRS relies on empirical calibration models, usually built with regression methods that link spectral measurements to laboratory reference data. Once suitable calibration models are available for multiple soil properties, they can be applied to a new soil spectrum to predict those properties simultaneously.

DRS methods used in soil science are commonly grouped into visible–near infrared (vis–NIR), shortwave infrared (SWIR), and mid-infrared (MIR) regions. The vis–NIR and SWIR regions are used in both laboratory and field applications, supported by the availability of portable instruments. These regions show a high sensitivity to soil organic matter, moisture, iron oxides, and texture-related features. MIR spectroscopy is more commonly conducted under laboratory conditions and often provides sharper absorption features associated with mineralogical and chemical soil attributes. In all these regions, spectral measurements of a soil sample can typically be acquired within seconds to a few minutes.

There is broad consensus that DRS complements rather than replaces soil laboratory analyses [@viscarra2022diffuse, @chen2023integrating], increasing analytical throughput while retaining the validity of reference procedures. For DSM, this means that spectroscopy can expand soil property datasets and improve mapping workflows while reducing cost and logistical constraints. This is true when the empirical models that translate spectra into soil property predictions have been properly calibrated and validated. It is important to note, however, that developing these models can still be laborious, time-consuming, and costly, a limitation that is rarely acknowledged in the soil spectroscopy literature. In this respect, pre-existing soil spectral libraries containing reference samples offer an opportunity to reduce calibration effort when developing new models for a specific area or application.

> _DRS can be a time- and cost-effective tool for soil analysis, as numerous studies have reported. However, this is mainly true when suitable predictive models have already been calibrated and validated. When models need to be developed from scratch, the process of generating reference data, calibrating robust models, and validating their performance can itself be costly and time-consuming. This limitation is often insufficiently acknowledged in the soil spectroscopy literature [@RamirezLopez2026Liblex]._


### Rationale 

The use of DRS-inferred soil properties offers a cost- and time-efficient pathway to substantially increase the number of observations available for DSM model calibration [@ramirez2019robust]. This directly supports the collection of spatially dense datasets at field to regional scales. By augmenting conventional laboratory datasets with larger volumes of DRS-inferred observations, DSM workflows can benefit from more input data, which may improve map accuracy at lower overall cost.

### Including Spectroscopic Predictions in Digital Soil Mapping

In DSM workflows that incorporate DRS, spectroscopically derived soil property estimates should be treated as model-based predictions with associated uncertainty, rather than as error-free measurements. Unlike conventional laboratory analyses, spectroscopy does not directly measure soil properties; it generates predictions from statistical models calibrated to capture relationships between spectral responses and reference measurements. Consequently, each DRS estimate is associated with uncertainty that is usually larger than that of conventional laboratory measurements. When DRS-inferred data are incorporated into DSM workflows, this additional source of uncertainty should therefore be explicitly acknowledged [@takoutsing2022accounting]. Treating DRS predictions as error-free observations may lead to biased spatial models, propagation of systematic spectroscopic errors, and overconfident DSM predictions [@van2022measurement].

Different methodological strategies have been proposed to propagate or account for uncertainty in soil property observations used in DSM, depending on the modeling framework adopted. In geostatistical mapping, observation-level uncertainty in soil property data has been incorporated into spatial covariance structures [@ramirez2019robust; @takoutsing2022accounting]. In machine-learning-based DSM, the literature remains more limited, although several studies have explored weighting schemes to reflect differences in data quality. For example, weights derived from DRS prediction uncertainty have been used in convolutional neural networks and random forest models by modifying the loss function during model calibration [@wadoux2019multi; @hengl2018random]. More recently, error-filtered machine learning frameworks have been proposed in which observation-level uncertainty is explicitly integrated into likelihood-based estimation, with improved predictive performance when observation errors are substantial [@van2022measurement].

A prerequisite for accounting for uncertainty in DRS predictions within DSM is that each observation includes an associated uncertainty estimate. In practice, DRS should provide not only a predicted soil property value, such as soil organic carbon content, but also an uncertainty metric, typically expressed as a prediction standard deviation or variance.

This is illustrated in Figure \@ref(fig:fig-uncertainty), which provides a conceptual representation of soil property observations obtained using conventional laboratory analysis and DRS. The laboratory measurement is shown as a narrow probability distribution centered on the reference value, reflecting relatively low analytical uncertainty. In contrast, the DRS-inferred value is represented by a wider probability distribution centered on the model prediction, reflecting larger prediction uncertainty. The dashed vertical lines denote the corresponding point estimates, while the horizontal arrows represent their standard deviations, $\sigma_{\mathrm{lab}}$ and $\sigma_{\mathrm{spectroscopy}}$. The displacement between the two distributions illustrates that spectroscopic predictions may differ systematically from reference measurements, introducing prediction bias in addition to random uncertainty.

DRS predictions of soil properties are generally less accurate than conventional laboratory measurements. However, their lower cost and higher analytical throughput can enable the acquisition of spatially denser datasets, which may improve the characterization of spatial variation and soil–environment relationships, provided that the associated prediction uncertainty is explicitly considered. In the practical section that follows, we illustrate this concept using an example of DSM implemented with a random forest model.

```{r fig-uncertainty, echo = FALSE, eval = TRUE, fig.cap="Conceptual illustration of measurement and prediction uncertainty. Narrow distributions represent wet-chemistry observations with lower analytical uncertainty, whereas wider distributions represent DRS predictions with larger associated standard deviations. Horizontal arrows indicate one standard deviation around the estimated value.", fig.width=6.5, fig.height=4.5}

library(ggplot2)

wet_mean <- 18.5
wet_sd <- 0.4
spec_mean <- 17.8
spec_sd <- 1.2

x <- seq(
  min(wet_mean - 4*wet_sd, spec_mean - 4*spec_sd),
  max(wet_mean + 4*wet_sd, spec_mean + 4*spec_sd),
  length.out = 1000
)

col_wet <- "#4E84A8" # steel blue
col_spec <- "#C1614A" # muted terracotta

df <- rbind(
  data.frame(x = x, density = dnorm(x, wet_mean, wet_sd), method = "Wet chemistry"),
  data.frame(x = x, density = dnorm(x, spec_mean, spec_sd), method = "DRS predictions")
)

means_df <- data.frame(
  mean = c(wet_mean, spec_mean),
  sd = c(wet_sd, spec_sd),
  method = c("Wet chemistry", "DRS predictions")
)

ymax <- max(df$density)

ggplot(df, aes(x = x, y = density, colour = method, fill = method)) +
  
  geom_rect(
    data = means_df,
    aes(xmin = mean - sd, xmax = mean + sd, ymin = 0, ymax = Inf, fill = method),
    alpha = 0.12, inherit.aes = FALSE
  ) +
  
  geom_area(alpha = 0.20, position = "identity") +
  geom_line(linewidth = 1.0) +
  
  geom_vline(
    data = means_df,
    aes(xintercept = mean, colour = method),
    linetype = "dashed", linewidth = 0.7, show.legend = FALSE
  ) +
  
  # sigma_DRS arrow — placed low, well left of the wet-chem peak
  annotate(
    "segment",
    x = spec_mean - spec_sd, xend = spec_mean + spec_sd,
    y = ymax * 0.38, yend = ymax * 0.38,
    colour = col_spec, linewidth = 0.8,
    arrow = arrow(ends = "both", length = unit(0.12, "cm"), type = "closed")
  ) +
  annotate(
    "text",
    x = spec_mean - spec_sd - 0.15, y = ymax * 0.38,
    label = expression(sigma[DRS]),
    colour = col_spec, size = 4.2, hjust = 1
  ) +
  
  # sigma_wet arrow — placed high, anchored right of its own peak
  annotate(
    "segment",
    x = wet_mean - wet_sd, xend = wet_mean + wet_sd,
    y = ymax * 0.72, yend = ymax * 0.72,
    colour = col_wet, linewidth = 0.8,
    arrow = arrow(ends = "both", length = unit(0.12, "cm"), type = "closed")
  ) +
  annotate(
    "text",
    x = wet_mean + wet_sd + 0.15, y = ymax * 0.72,
    label = expression(sigma["wet chem."]),
    colour = col_wet, size = 4.2, hjust = 0
  ) +
  
  scale_colour_manual(
    values = c("Wet chemistry" = col_wet, "DRS predictions" = col_spec)
  ) +
  scale_fill_manual(
    values = c("Wet chemistry" = col_wet, "DRS predictions" = col_spec)
  ) +
  
  labs(
    x = "Soil property value", 
    y = "Probability density", 
    colour = NULL, 
    fill = NULL
  ) +
  
  theme_classic(base_size = 13) +
  theme(
    legend.position = "top",
    legend.key.size = unit(0.9, "lines"),
    legend.text = element_text(size = 11),
    axis.line = element_line(colour = "grey40"),
    panel.grid.major.y = element_line(colour = "grey92", linewidth = 0.4), 
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.border = element_rect(colour = "grey40", fill = NA, linewidth = 0.5)
  )
```



## Workflow overview  

In this practical section, we illustrate how DRS-derived soil property estimates can complement conventional laboratory analyses in the production of digital soil maps. The methodology is demonstrated using the Kansas, USA, dataset introduced in Chapter \@ref(digital-soil-mapping). In addition to the soil property variables used in the DSM example, this dataset includes mid-infrared DRS measurements that were not used previously.

Figure \@ref(fig:fig-dsm-workflow) presents the conceptual workflow used in this section. The colors distinguish the main groups of steps:

- <span style="color:#4E84A8; font-weight:700;">Blue boxes</span> represent field sampling and laboratory reference analysis. These steps provide the measured soil property data used as calibration targets, either to fit a DRS calibration model or to support DSM.

- <span style="color:#C1614A; font-weight:700;">Terracotta boxes</span> correspond to the DRS component. This includes spectral acquisition, calibration modeling, and the generation of spectroscopically predicted soil properties with associated uncertainty, expressed as the prediction error standard deviation. These predictions are treated as an additional source of soil information in the DSM step.

- <span style="color:#7B6D9E; font-weight:700;">Purple boxes</span> indicate the data integration stage. Here, laboratory observations and DRS predictions are combined into a single dataset.

- <span style="color:#4E8A6A; font-weight:700;">Green boxes</span> represent the spatial modeling stage, or DSM. Environmental covariates are used together with soil observations to fit spatial prediction models. The uncertainty associated with spectroscopic predictions is used to down-weight these observations relative to laboratory measurements. The final outputs are continuous soil property maps and corresponding uncertainty maps.

All main steps in this workflow are implemented and illustrated using R in the following subsections.

```{r fig-dsm-workflow, echo=FALSE, message=FALSE, warning=FALSE, out.width="80%", fig.cap="Conceptual workflow for integrating spectroscopically predicted soil data into digital soil mapping.", eval=TRUE}
library(DiagrammeR)
library(DiagrammeRsvg)
library(rsvg)

g <- grViz("
digraph dsm_workflow {

  graph [layout = dot, rankdir = TB, nodesep = 0.3, ranksep = 0.2
         bgcolor = '#FFFFFF']

  node [shape = box,
        style = 'rounded,filled',
        penwidth = 1.5,
        fontname = 'Helvetica Neue',
        fontsize = 10,
        margin = 0.2]

  edge [arrowsize = 0.7,
        penwidth = 1.2,
        color = '#B0B8C1']

  sampling [label = 'Soil sampling',                              fillcolor = '#D6E8F2', color = '#4E84A8', fontcolor = '#1A3D52']
  lab      [label = 'Lab analysis',                               fillcolor = '#D6E8F2', color = '#4E84A8', fontcolor = '#1A3D52']
  spectra  [label = 'Spectral scanning',                          fillcolor = '#F2DDD9', color = '#C1614A', fontcolor = '#5C2518']
  calib    [label = 'Calibration model',                          fillcolor = '#F2DDD9', color = '#C1614A', fontcolor = '#5C2518']
  pred     [label = 'Spectroscopic predictions\\n(+ uncertainty)', fillcolor = '#F2DDD9', color = '#C1614A', fontcolor = '#5C2518']
  harmon   [label = 'Data harmonisation\\n/ weighting',           fillcolor = '#E2DDED', color = '#7B6D9E', fontcolor = '#312650']
  covars   [label = 'Environmental covariates',                   fillcolor = '#D8E8DF', color = '#4E8A6A', fontcolor = '#1A3D2A']
  dsm      [label = 'DSM model',                                  fillcolor = '#D8E8DF', color = '#4E8A6A', fontcolor = '#1A3D2A']
  maps     [label = 'Predicted soil maps',                        fillcolor = '#D8E8DF', color = '#4E8A6A', fontcolor = '#1A3D2A']

  sampling -> lab
  sampling -> spectra

  lab -> calib
  spectra -> calib

  calib -> pred
  lab -> harmon
  pred -> harmon

  covars -> dsm
  harmon -> dsm

  dsm -> maps
}
")
out_png <- "images/module4/fig_dsm_workflow.png"
tmp_svg <- tempfile(fileext = ".svg")
export_svg(g) |> writeLines(tmp_svg)
rsvg_png(tmp_svg, out_png, width = 1400)
knitr::include_graphics(out_png)
```

## Generating DRS-derived soil property observations for DSM

In this section, we convert DRS measurements into soil property estimates that can be used as additional observations in DSM. The key requirement is that, for each soil sample, the spectroscopic modeling step provides both a predicted value of the target soil property and an estimate of prediction uncertainty. This follows the approach introduced in the Soil Spectroscopy Training Material (@Wadoux2025FAO, Section 4.3.3), where uncertainty is quantified at the sample level.

For DSM applications, these outputs can be treated similarly to laboratory observations. However, unlike laboratory data, DRS predictions are associated with explicit model-based uncertainty. This uncertainty information can later be used during data integration and spatial modeling.

In the following subsections, we demonstrate how to (i) import and visualize the spectral data, (ii) apply basic spectral preprocessing, (iii) build a spectroscopic prediction model with uncertainty, (iv) define laboratory measurement uncertainty, and (v) harmonize soil property predictions and uncertainties to a standard depth interval.

### Load and plot the DRS data

First, we import the DRS dataset and prepare it for analysis. This dataset contains absorbance measurements collected in the mid-infrared (MIR) region of the electromagnetic spectrum, over the wavenumber range starting from apprximately 4000 cm$^{-1}$ up to apprximately 400 cm$^{-1}$ at an approximated resolution of 1.93 cm$^{-1}$. This resulted in 1764 spectral variables. Note that this region corresponds to wavelengths of approximately 2500–25000 nm. These measurements were collected under laboratory conditions using air-dried soil samples. Additional details on the DRS measurements are provided in @dangal2019accurate.


> <span  style="font-weight:700;">NOTE:</span> _MIR spectra show strong, well-defined absorption features associated with molecular vibrations in soil minerals and organic compounds. This makes MIR spectroscopy sensitive to properties such as soil organic carbon, clay minerals, carbonates, and some nutrients.
Although this example uses MIR spectra, the workflow can also be applied to vis–NIR or SWIR data, provided suitable calibration models and uncertainty estimates are available._


Let's start by loading the data:

```{r message=FALSE, warning=FALSE, eval=TRUE}
library(readxl)
library(dplyr)

# Read MIR spectral dataset (Kansas example)
original_dat <- read_excel(
  "01_data/module1/training_data/data/MIR_KANSAS_data.xlsx",
  .name_repair = "minimal",
  na = c("", "NA", "N/A", "NaN"),
  progress = FALSE
)

original_dat <- as.data.frame(original_dat)

# dimensions of the dataset (number of samples and variables)
dim(original_dat)

# number of duplicate sample IDs (i.e. repeated measurements for 
# the same soil sample)a
sum(duplicated(original_dat$smp_id))
```
The loaded dataset contains both DRS variables and non-DRS variables. The first step is therefore to separate the spectral data from the rest of the soil dataset. To do this, we create a data frame containing the non-spectral variables and store the DRS variables as a spectral matrix. This structure allows the spectra to be handled separately for preprocessing and modeling, while keeping them linked to the corresponding soil samples.

The spectral variables can be identified from the column names. These can be inspected using `colnames(dat)`. In this dataset, spectral columns follow a consistent naming pattern: they start with the prefix `X`, followed by numeric values representing the original wavenumbers. For example, a column named `X3999.7` corresponds to a spectral variable measured at approximately 3999.7 cm$^{-1}$.

We use this pattern to separate the spectral variables from the remaining soil data. The `X` prefix is then removed so that the column names can be converted into numeric wavenumber values. These values are stored in `original_wavs` and used later for plotting, resampling, and spectral preprocessing.

```{r message=FALSE, warning=FALSE, eval=TRUE}
# Identify spectral columns
# Spectral variables start with "X" followed by at least three digits.
spc_cols <- grep("^X[0-9]{3,}", colnames(original_dat))

# Extract the spectral matrix
spc_original <- original_dat[, spc_cols]

# Remove the "X" prefix from spectral column names
colnames(spc_original) <- gsub("^X", "", colnames(spc_original))

# Convert spectral column names to numeric wavenumber values
original_wavs <- as.numeric(colnames(spc_original))

# Check the spectral range
range(original_wavs)
```

Because the spectra are recorded at high spectral resolution, the spectral matrix contains many highly correlated variables. Before modeling, we therefore resample the spectra to a coarser wavenumber grid. This reduces computational demand while preserving the main absorption features. In this case, we resample the spectra to a resolution of 8 cm$^{-1}$ using `prospectr::resample()`. Since the original spectra do not extend exactly to 4000 cm$^{-1}$, we avoid extrapolation (highly recommended for all cases) and resample only up to 3992 cm$^{-1}$. This results in a new spectral matrix with 425 spectral variables, which is more manageable for modeling while still capturing the key spectral information:

```{r message=FALSE, warning=FALSE, eval=TRUE}
library(prospectr)
# Define a new regular spectral grid for resampling
wavs <- seq(600, 3992, by = 8)
# Resample spectra to the new wavelength grid
spc_resampled <- prospectr::resample(
  spc_original,
  wav = original_wavs,
  new.wav = wavs
)
# check the dimensions
dim(spc_resampled)
```

Now we can plug this resampled spectral matrix into the non-DRS part of the dataframe:
```{r message=FALSE, warning=FALSE, eval=TRUE, results='hide'}
# extract non-spectral variables into a new dataframe
repeated_dat <- original_dat[, -spc_cols]

# in the main dataframe create a new variable called spc 
# containing the matrix of spectra 
repeated_dat$spc <- spc_resampled

# make some checks
colnames(repeated_dat)
dim(repeated_dat)
dim(repeated_dat$spc)
```

The data contains repeated measurements for each soil sample, which is common in spectroscopic datasets to account for measurement variability. To simplify the analysis, we start by averaging replicated spectral measurements so that each soil sample is represented by a single spectrum.

```{r message=FALSE, warning=FALSE, eval=TRUE, results = 'hide'}
# Average replicate spectra by sample ID
spc_avg <- aggregate(
  repeated_dat$spc,
  by = list(smp_id = repeated_dat$smp_id),
  FUN = mean
)
# order the resuslts by sample ID to ensure correct alignment with the main 
# dataset later on
spc_avg <- spc_avg[order(spc_avg$smp_id),  ]

# aggregate() prepends smp_id as first column; extract the spectral matrix
rownames(spc_avg) <- spc_avg$smp_id
spc_avg <- as.matrix(spc_avg[, -1])

# Drop duplicate non-spectral rows (keep first occurrence per sample)
dat <- repeated_dat[!duplicated(repeated_dat$smp_id), ]
dat <- dat[order(dat$smp_id), ]

# Verify that sample IDs are aligned before overwriting spectra
stopifnot(all(rownames(spc_avg) == dat$smp_id))

# Overwrite spectra with the averaged matrix
dat$spc <- spc_avg

# check dimensions of the final dataset
dim(dat)
```

Now let's plot the spectra (Figure \@ref(fig:fig-spectra)) that we have so far:
```{r fig-spectra, message=FALSE, warning=FALSE, eval=TRUE, out.width="80%", fig.cap="Resampled and aggreagted MIR sectra."}
matplot(
  wavs,
  t(dat$spc),
  type = "l",
  lty = 1,
  xlim = c(4000, 500),
  xlab = expression(Wavenumber~(cm^{-1})),
  ylab = "Absorbance",
  col = rgb(0.8, 0, 0, 0.3)
)
grid(lty = 1)
```

A unique profile identifier is also created to group horizons belonging to the same soil profile:

```{r message=FALSE, warning=FALSE, eval=TRUE, results = 'hide'}
# Create unique profile IDs for each coordinate pair
coords <- paste(dat$Long_Site.x, dat$Lat_Site.x, sep = "_")
prof_num <- match(coords, unique(coords))

dat$ProfID <- sprintf("PROF%04d", prof_num)
```

### Preprocessing of the soil spectra

Raw soil spectra often require preprocessing before modeling. The objective is to reduce instrumental or measurement artifacts while preserving the chemical information relevant for prediction.

A first step is always to inspect the spectra visually. This helps identify evident noise, baseline shifts, outliers, or unstable regions at the edges of the spectral range. Edge noise can occur because DRS detectors often have lower sensitivity near the limits of their operating range. When this occurs, it is common practice to trim the affected regions before modeling. In this example, we do not trim the spectra, but this should be considered for other datasets.

Spectra can also be affected by variation in instrument response, sample presentation, particle size, packing density, and measurement conditions. These effects may introduce additive or multiplicative changes in the spectra, even when the same sample is measured more than once. Common preprocessing methods used to reduce these effects include:

- **Smoothing**, such as moving average or Savitzky–Golay filtering, to reduce high-frequency noise.
- **Scatter correction**, such as standard normal variate (SNV) or multiplicative scatter correction (MSC), to reduce multiplicative effects.
- **Derivatives**, often combined with smoothing, to reduce baseline effects and enhance spectral features.

Derivatives should be used with caution because they can also amplify noise. The choice of preprocessing method should therefore depend on the spectral region, the quality of the measurements, and the modeling objective.

> <span  style="font-weight:700;">NOTE:</span> _Scatter correction methods may sometimes appear to reduce model accuracy during internal validation, especially when calibration and validation samples come from the same instrument, preparation protocol, and measurement conditions. In such cases, raw or lightly smoothed spectra may preserve dataset-specific variation that improves apparent performance. However, this variation may not be chemically meaningful and can make the model more sensitive to changes in particle size, packing density, sample presentation, instrument response, or acquisition conditions. Carefully selected scatter correction methods, such as SNV or MSC, may therefore produce models that are slightly less accurate in the short term but more robust when applied to new samples, instruments, or measurement settings. For this reason, scatter correction should be evaluated not only in terms of cross-validation accuracy, but also in terms of expected model transferability and long-term stability._

In this example, we first apply a Savitzky–Golay first derivative to reduce baseline effects and enhance spectral features, followed by standard normal variate (SNV) correction to reduce multiplicative scatter effects. In `savitzkyGolay()`, `m = 1` specifies the first derivative, `p = 1` sets the polynomial order used within the moving window, and `w = 25` defines the window size in spectral variables. The optimal preprocessing strategy should, however, be evaluated for each dataset.

```{r message=FALSE, warning=FALSE, eval =TRUE}
# apply standard normal variate (SNV) correction followed by 
# a Savitzky–Golay first derivative
dat$spc_processed  <- dat$spc |> 
  prospectr::savitzkyGolay(m = 1, p = 1, w = 25) |> 
  prospectr::standardNormalVariate() 

# get the wavenumber values for the processed spectra
wavs_processed <- as.numeric(colnames(dat$spc_processed))
```


Now let's visualise the processed spectra(Figure \@ref(fig:fig-spectra-processed)) that we have so far:
```{r fig-spectra-processed, message=FALSE, warning=FALSE, eval=TRUE, out.width="80%", fig.cap = "Processed MIR spectra after standard normal variate correction and Savitzky–Golay first-derivative transformation."}
matplot(
  wavs_processed,
  t(dat$spc_processed),
  type = "l",
  lty = 1,
  xlim = c(4000, 500),
  xlab = expression(Wavenumber~(cm^{-1})),
  ylab = "SNV of first derivative absorbance", 
  col = rgb(0.3, 0, 0.8, 0.3)
)
grid(lty = 1)
```


This basic exploratory step helps verify that the spectra have been imported and processed correctly. In a full analysis, this stage can also be used to screen for unusual spectra or potential measurement issues before fitting the calibration model. The dataset is now ready for statistical modeling.


### SOC prediction from DRS spectra with uncertainty estimation

#### Select a subset of calibration samples for model fitting

Here, the objective is to use a small but representative subset of samples to calibrate a DRS model for predicting SOC. The remaining samples are then treated as if their SOC content were unknown and are predicted using the calibrated DRS model.

This setup simulates a practical scenario in which only a limited number of samples are analysed using conventional laboratory methods, while the remaining samples are characterised through DRS-based predictions. This is particularly relevant for digital soil mapping (DSM), where the aim is often to increase the spatial density of soil property information by combining a smaller set of laboratory measurements with a larger number of spectroscopic predictions.

> _The rationale is to simulate a cost-efficient scenario in which conventional laboratory analysis is performed only on a limited, representative subset of samples, while DRS is used to extend soil property information to a larger number of samples through model-based prediction._

Here, we assume that approximately 20% of the dataset is sufficient to calibrate a robust DRS model for SOC prediction. The calibration subset was selected using the Kennard–Stone algorithm, implemented in the `kenStone()` function of the `prospectr` package. This calibration sampling method selects observations that are well distributed across the spectral space, so that the selected subset captures the main spectral variability present in the full dataset.

The selection IS based on the processed DRS spectra. Principal component analysis is performed internally by `kenStone()`, and here we specified that the first 10 principal components should be used. Samples were then selected in this principal component space using Mahalanobis distance as the distance metric. A total of 517 samples was targeted for calibration. The selection was constrained by soil profile identity (`ProfID`). Instead of selecting individual samples independently, profiles were selected as groups, and all samples from the selected profiles were used for model calibration. This avoids pseudo-replication and potential information leakage, since samples from the same soil profile are not fully independent. Keeping profiles intact provides a more realistic separation between calibration and prediction samples.

```{r message=FALSE, warning=FALSE, eval=TRUE}
# Number of samples targeted for the calibration subset
n_cal_samples <- 517

# Select a spectrally representative calibration subset using Kennard-Stone.
# The selection is based on Mahalanobis distances computed in the space of the
# first 10 principal components of the processed spectra.
#
# The grouping variable ProfID ensures that selection is performed at the
# soil-profile level rather than at the individual-sample level. All samples
# from selected profiles are used for calibration, avoiding pseudo-replication
# and information leakage between calibration and prediction subsets.
ks_samples <- prospectr::kenStone(
  dat$spc_processed, 
  k = n_cal_samples, 
  metric = "mahal", 
  pc = 10, 
  group = as.factor(dat$ProfID)
)
```

The `ks_samples$model` vector contains the indices of the samples selected for model calibration. The remaining samples are treated as if their SOC content were unknown and are predicted using the calibrated DRS model. So we split the data into calibration and prediction subsets:

```{r message=FALSE, warning=FALSE, eval=TRUE}
dat_cal <- dat[ks_samples$model, ]
dat_pred <- dat[-ks_samples$model, ]
```

#### Model calibration

In this section, we calibrate a random forest (RF) model to predict SOC from DRS spectra. The RF algorithm used is the _quantile regression forest_, which provides a prediction distribution for each sample rather than only a single point estimate. This allows us to obtain both the predicted SOC value and an associated uncertainty estimate, expressed here as the prediction standard deviation. 

For the RF modeling setp, we use the `ranger` package [@wright2017ranger].

Prediction uncertainty is quantified by the model using the conditional distribution of the predictions. This approach is described in detail for soil DRS data by @wadoux2025uncertainty. From this model distribution, a mean predicted value, a prediction variance, and a prediction standard deviation are calculated for each soil sample. These quantities represent the required outputs of the DRS modeling step for DSM.

To obtain an assessment of model performance, cross-validation is done at the soil profile level. This avoids using horizons from the same profile in both training and testing subsets. For each validation fold, the model is trained on a subset of profiles and used to predict SOC for the remaining profiles.

Model performance is evaluated using cross-validation at the soil profile level. All horizons from the same profile are assigned to the same fold, so that horizons from one profile are never split between calibration and validation subsets. In each iteration, one fold of profiles is left out, the RF model is calibrated using the remaining profiles, and SOC is predicted for the excluded profiles. This avoids pseudo-replication and provides a more realistic assessment of model performance. The predicted SOC values are then compared with the laboratory reference values to compute performance metrics such as bias, RMSE, and $R^2$. The validation results are visualized by plotting predicted SOC values against the corresponding laboratory reference measurements, with error bars indicating the prediction standard deviation for each sample.

Let's first define the fold structure for cross-validation. We assign folds at the profile level to ensure that all horizons from the same soil profile are kept together in either the training or test set, preventing information leakage and providing a more realistic evaluation of model performance.

```{r message=FALSE, warning=FALSE, eval=TRUE}
# set.seed() makes the random assignment reproducible: rerunning the
# script always produces the same fold structure. The value 123 is arbitrary.
set.seed(14092019) # just any number

# 10 folds
nfolds <- 10
# We assign folds at the profile level because observations within a
# profile are depth-wise measurements of the same soil column... they
# are not independent. Splitting at the observation level would leak
# information from the same profile into both training and test sets.
cal_profile_ids <- unique(dat_cal$ProfID)
fold_index  <- sample(rep(1:nfolds, length.out = length(cal_profile_ids)))

# Each observation inherits the fold of its parent profile
dat_cal$fold <- fold_index[match(dat_cal$ProfID, cal_profile_ids)]
```

Now we can proceed with the modeling. In this implementation  for each fold we use the option `quantreg = TRUE` which enables estimation of the conditional distribution of SOC rather than only a single point prediction. The forest was grown with 3500 trees, bootstrap sampling with replacement, and a minimum node size of 10. Splits were selected using the maximally selected rank statistic (`splitrule = "maxstat"`), and a fixed seed was used for reproducibility. For each held-out sample, 100 values were drawn from the estimated conditional distribution and summarised by their mean, variance, and standard deviation. Further details on the `ranger` implementation and this uncertainty-aware modelling approach are provided by @wright2017ranger and @wadoux2025uncertainty.

```{r message=FALSE, warning=FALSE, eval=TRUE, results = 'hide'}
# Load the ranger package for quantile forest modeling 
library(ranger)
library(matrixStats)

# Prepare storage vectors for CV predictions and uncertainty
SOC_predRF <- rep(NA_real_, nrow(dat_cal))  # mean prediction
SOC_varRF <- rep(NA_real_, nrow(dat_cal))  # predictive variance
SOC_sdRF <- rep(NA_real_, nrow(dat_cal))  # predictive standard deviation

# Run 10-fold cross-validation
for (k in seq_len(nfolds)) {
  cat("\rProcessing fold", k, "out of 10")
  # Test set: all samples belonging to fold k
  test_rows <- which(dat_cal$fold == k)
  # Training set: all remaining samples
  train_rows <- which(dat_cal$fold != k)
  # Fit quantile random forest on training data
  model <- ranger(
    x = dat_cal$spc_processed[train_rows, ],
    y = dat_cal$SOC[train_rows],
    quantreg = TRUE,
    num.trees = 3500,
    sample.fraction = 1,
    replace = TRUE,
    splitrule = "maxstat",
    min.node.size = 10, 
    seed = 201909
  )
  
  # Predict by drawing samples from the conditional distribution
  pred_matrix <- predict(
    model,
    data = dat_cal$spc_processed[test_rows, ],
    type = "quantiles",
    what = function(x) sample(x, 100, replace = TRUE)
  )$predictions
  
  # Store prediction summary statistics
  SOC_predRF[test_rows] <- rowMeans(pred_matrix)
  SOC_varRF[test_rows] <- rowVars(pred_matrix)
  SOC_sdRF[test_rows] <- sqrt(SOC_varRF[test_rows])
}

# Save predictions back into the dataset
dat_cal$SOC_predRF <- SOC_predRF
dat_cal$SOC_varRF <- SOC_varRF
dat_cal$SOC_sdRF <- SOC_sdRF
```

Figure \@ref(fig:fig-val-plot) shows the validation plot of predicted SOC values against laboratory reference measurements. Error bars represent the prediction standard deviation for each sample, providing a visual representation of the uncertainty associated with each prediction. 

```{r echo=TRUE, message=FALSE, warning=FALSE, eval=FALSE}
xy_lims <- range(dat_cal$SOC_predRF, dat_cal$SOC)
# Validation plot
plot(
  x = dat_cal$SOC_predRF,
  y = dat_cal$SOC,
  xlab = "Predicted SOC (%)",
  ylab = "Measured SOC (%)",
  xlim = xy_lims,
  ylim = xy_lims,
  col = rgb(0, 0, 0, 0.5), 
  pch = 16, 
  cex = 1.5
)

# add the error bars representing the prediction standard deviation
arrows(
  x0 = dat_cal$SOC_predRF - dat_cal$SOC_sdRF,
  y0 = dat_cal$SOC,
  x1 = dat_cal$SOC_predRF + dat_cal$SOC_sdRF,
  y1 = dat_cal$SOC,
  code = 3,
  length = 0,
  col = rgb(0, 0, 0, 0.5)
)

abline(0, 1, col = col_ref, lwd = 1.5, lty = 2)
grid(col = rgb(0.8, 0.8, 0.8, 0.6), lty = 1)
```

```{r fig:fig-val-plot, echo=FALSE, message=FALSE, warning=FALSE, eval=TRUE,  fig.cap="Validation plot of predicted SOC values against laboratory reference measurements. Error bars represent the prediction standard deviation for each sample. The dashed line indicates the 1:1 relationship, which represents perfect agreement between predictions and measurements."}
col_errbar <- rgb(0.20, 0.45, 0.70, 0.35) # steel blue, low alpha
col_points <- rgb(0.15, 0.15, 0.15, 0.75) # near-black
col_ref <- rgb(0.50, 0.50, 0.50, 0.90) # mid grey

xy_lims <- range(dat_cal$SOC_predRF, dat_cal$SOC)
plot(
  x = xy_lims,
  y = xy_lims,
  xlab = "Predicted SOC (%)",
  ylab = "Measured SOC (%)",
  xlim = xy_lims,
  ylim = xy_lims,
  pch = 16,
  col = NA
)
arrows(
  x0 = dat_cal$SOC_predRF - dat_cal$SOC_sdRF,
  y0 = dat_cal$SOC,
  x1 = dat_cal$SOC_predRF + dat_cal$SOC_sdRF,
  y1 = dat_cal$SOC,
  code = 3,
  length = 0,
  col = col_errbar
)
points(
  x = dat_cal$SOC_predRF,
  y = dat_cal$SOC,
  pch = 16,
  col = col_points
)
abline(0, 1, col = col_ref, lwd = 1.5, lty = 2)
grid(col = rgb(0.8, 0.8, 0.8, 0.6), lty = 1)
```

The root mean squared error (RMSE) of such type of quantification models can be computed as follows:
```{r echo=TRUE, message=FALSE, warning=FALSE, eval=TRUE}
drs_rmse <- sqrt(mean((dat_cal$SOC_predRF - dat_cal$SOC)^2, na.rm = TRUE))
cat("DRS model RMSE:", round(drs_rmse, 3), "% SOC\n")
```

The final model is fitted using all the calibration data at once:

```{r message=FALSE, warning=FALSE, eval=TRUE}
final_drs_model <- ranger(
  x = dat_cal$spc_processed,
  y = dat_cal$SOC,
  quantreg = TRUE,
  num.trees = 3500,
  sample.fraction = 1,
  replace = TRUE,
  splitrule = "maxstat",
  min.node.size = 10, 
  seed = 201909
)
```


#### DRS-based prediction and uncertainty estimation for non-calibration samples

The prediction in the remaining 80% of the samples is done by applying the final DRS model to the processed spectra of the prediction subset. The same approach is used to obtain both predicted SOC values and associated uncertainty estimates for these samples:

```{r message=FALSE, warning=FALSE, eval=TRUE}
# Predict for the remaining samples using the final model
drs_soc_preds <- predict(
  final_drs_model,
  data = dat_pred$spc_processed,
  type = "quantiles",
  what = function(x) sample(x, 100, replace = TRUE)
)$predictions
  
# Store prediction summary statistics
dat_pred$SOC_predRF <- rowMeans(drs_soc_preds)
dat_pred$SOC_varRF <- rowVars(drs_soc_preds)
dat_pred$SOC_sdRF <- sqrt(dat_pred$SOC_varRF)
```

## Assembling the augmented dataset of laboratory measurements and DRS-based predictions

For DSM, we assemble a final spatial modelling dataset for SOC using two sources of information:

- **Laboratory SOC measurements** for the samples used to calibrate the DRS model. These samples represent approximately 20% of the original dataset. Because these values are assumed to be available from conventional laboratory analysis, they are retained in preference to DRS-based predictions.

- **DRS-based SOC predictions** for the remaining samples, which were not used in the calibration of the DRS model. As described above, these samples are treated as if laboratory SOC values were unavailable. Their SOC values are therefore estimated entirely from the DRS model. This setup simulates a realistic application scenario in which DRS is used to increase the number of SOC observations available for DSM.

```{r message=FALSE, warning=FALSE, eval=TRUE}
# Extract metadata for the augmented dataset
dat_augmented <- dat[, c(
  "smp_id", "ProfID", "Long_Site.x", "Lat_Site.x", 
  "Top_depth_cm.x", "Bottom_depth_cm.x"
  )
]

# Initialise SOC values and associated standard deviations
dat_augmented$SOC <- NA
dat_augmented$SOC_sd <- NA

# Add laboratory SOC values for the samples used to calibrate the DRS model
dat_augmented$SOC[ks_samples$model] <- dat$SOC[ks_samples$model]

# Add DRS-based SOC predictions for the remaining samples
dat_augmented$SOC[-ks_samples$model] <- dat_pred$SOC_predRF 
```

Add the deviations from the DRS model as the uncertainty of the DRS-based predictions:
```{r message=FALSE, warning=FALSE, eval=TRUE}
dat_augmented$SOC_sd[-ks_samples$model] <- dat_pred$SOC_sdRF 
```
## Uncertainty of the conventional laboratory measurement

The SOC concentrations in the dataset used here were determined by high-temperature dry combustion, one of the most precise laboratory approaches for SOC quantification and the standard reference method for this property.

Quantifying laboratory measurement uncertainty in the absence of replicate measurements requires an external reference. @stevens2013prediction assessed the reproducibility of dry combustion SOC analysis across a large harmonised European soil survey and reported a standard error of laboratory (SEL) of approximately 2 g C kg⁻¹ for mineral soils overall — equivalent to 0.2% SOC. This value reflects intermediate precision under standardised conditions in an accredited laboratory and represents a realistic but optimistic estimate of the analytical component of uncertainty, excluding sampling and preparation variability.

For this dataset, which was similarly analysed under standardised laboratory conditions [@dangal2019accurate], we adopt a constant absolute uncertainty of $\sigma_{\text{lab}} = 0.15$% SOC as a pragmatic approximation consistent with the published literature. A constant rather than proportional error model is used here because a proportional formulation — where $\sigma = CV \times y_i$ — causes low-SOC observations to receive disproportionately high weights during model calibration, irrespective of their actual reliability. In practical applications, laboratory-specific replicate measurements should be used to derive empirical uncertainty estimates.

```{r message=FALSE, warning=FALSE, eval=TRUE}
# Constant absolute laboratory uncertainty: 0.15% SOC
# Based on Stevens et al. (2013), SEL ≈ 2 g C kg-1 for mineral soils
dat$SOC_sd <- 0.15
```

The same value is assigned to the laboratory observations in the augmented dataset:

```{r message=FALSE, warning=FALSE, eval=TRUE}
dat_augmented$SOC_sd[ks_samples$model] <- 0.15
```

## Depth harmonization of predictions and uncertainties

Digital soil mapping often requires soil properties to be expressed for standard depth intervals, such as 0–30 cm. However, field observations and DRS predictions are typically available at the horizon level, with horizons varying in thickness. In this step, SOC values and their associated uncertainties are aggregated to the 0–30 cm interval using depth-weighted averaging. Profiles that do not fully cover the target interval are excluded. This harmonization is applied to both the reference dataset, which contains laboratory measurements only, and the augmented dataset, which combines laboratory measurements with DRS-based predictions.

We first load the `aqp` package and convert both datasets to `SoilProfileCollection` objects, which store horizon-level data in a structured format that supports depth-based operations:

```{r dh-build-spc, message=FALSE, warning=FALSE, eval=TRUE}
library(aqp)

depths(dat) <- ProfID ~ Top_depth_cm.x + Bottom_depth_cm.x
depths(dat_augmented) <- ProfID ~ Top_depth_cm.x + Bottom_depth_cm.x
```

Combining uncertainty estimates across horizons requires variance propagation. For a depth-weighted average $\hat{y} = \sum_i w_i y_i / \sum_i w_i$, where $w_i$ is the thickness of horizon $i$ within the target interval, the aggregated standard deviation is $\sigma = \sqrt{\sum_i w_i^2 \sigma_i^2} / \sum_i w_i$. The helper function below implements this for an arbitrary depth interval, returning `NA` for profiles that do not fully cover it:

```{r dh-helper-fn, message=FALSE, warning=FALSE, eval=TRUE}
weighted_sd_0_30 <- function(top, bottom, sd, z1 = 0, z2 = 30) {
  w    <- pmax(0, pmin(bottom, z2) - pmax(top, z1))
  keep <- w > 0 & !is.na(sd)
  w    <- w[keep]
  sd   <- sd[keep]
  if (sum(w) < (z2 - z1)) return(NA)
  sqrt(sum((w * sd)^2)) / sum(w)
}
```

### Reference dataset

The reference dataset contains laboratory SOC measurements only. We compute the depth-weighted mean SOC and its associated uncertainty (approximated as 3% of the measured value, as defined earlier) for the 0–30 cm interval:

```{r dh-ref, message=FALSE, warning=FALSE, eval=TRUE}
soc_mean_ref <- profileApply(dat, function(p) {
  h <- horizons(p)
  w <- pmax(0, pmin(h$Bottom_depth_cm.x, 30) - pmax(h$Top_depth_cm.x, 0))
  keep <- w > 0 & !is.na(h$SOC)
  if (sum(w[keep]) < 30) return(NA)
  sum(w[keep] * h$SOC[keep]) / sum(w[keep])
})

soc_sd_ref <- horizons(dat) |>
  as.data.frame() |>
  group_by(ProfID) |>
  summarise(
    SOC_sd_0_30 = weighted_sd_0_30(Top_depth_cm.x, Bottom_depth_cm.x, SOC_sd),
    .groups = "drop"
  )

coords_ref <- as.data.frame(dat)[, c("ProfID", "Lat_Site.x", "Long_Site.x")]
coords_ref <- coords_ref[!duplicated(coords_ref$ProfID), ]

soc_ref_0_30_xy <- soc_sd_ref |>
  mutate(SOC_0_30 = soc_mean_ref) |>
  select(ProfID, SOC_0_30, SOC_sd_0_30) |>
  merge(coords_ref, by = "ProfID", all.x = TRUE)

head(soc_ref_0_30_xy)
```

### Augmented dataset

The augmented dataset combines laboratory measurements with DRS-based predictions. The single `SOC_sd` column carries the appropriate uncertainty for each observation — laboratory analytical error for reference samples, RF prediction uncertainty for DRS-predicted samples — so the same function applies without modification:

```{r dh-aug, message=FALSE, warning=FALSE, eval=TRUE}
soc_mean_aug <- profileApply(
  dat_augmented, 
  function(p) {
    h <- horizons(p)
    w <- pmax(0, pmin(h$Bottom_depth_cm.x, 30) - pmax(h$Top_depth_cm.x, 0))
    keep <- w > 0 & !is.na(h$SOC)
    if (sum(w[keep]) < 30) 
      return(NA)
    sum(w[keep] * h$SOC[keep]) / sum(w[keep])
  }
)

soc_sd_aug <- horizons(dat_augmented) |>
  as.data.frame() |>
  group_by(ProfID) |>
  summarise(
    SOC_sd_0_30 = weighted_sd_0_30(Top_depth_cm.x, Bottom_depth_cm.x, SOC_sd),
    .groups = "drop"
  )

coords_aug <- as.data.frame(dat_augmented)[, c("ProfID", "Lat_Site.x", "Long_Site.x")]
coords_aug <- coords_aug[!duplicated(coords_aug$ProfID), ]

soc_aug_0_30_xy <- soc_sd_aug |>
  mutate(SOC_0_30 = soc_mean_aug) |>
  select(ProfID, SOC_0_30, SOC_sd_0_30) |>
  merge(coords_aug, by = "ProfID", all.x = TRUE)

head(soc_aug_0_30_xy)
```
Figure \@ref(fig:fig-soc-distributions) illustrates the empirical probability distributions of SOC values in the reference and augmented datasets after depth harmonization. These two distributions are very similar and have comparable means and standard deviations, indicating that the DRS-based predictions are consistent with the laboratory measurements in terms of their central tendency and variability. 

```{r echo = TRUE, eval = FALSE}
dens_ref <- density(soc_ref_0_30_xy$SOC_0_30, na.rm = TRUE)
dens_aug <- density(soc_aug_0_30_xy$SOC_0_30, bw = dens_ref$bw, na.rm = TRUE)

plot(
  dens_ref$x, 
  dens_ref$y, 
  type = "l", 
  col = "#4E84A8", 
  lwd = 2, 
  ylab = "Density",
  xlab = "SOC, %"
)
lines(
  dens_aug$x, 
  dens_aug$y, 
  col = "#C1614A", 
  lwd = 2
)
legend(
  "topright",
  legend = c("Conventional lab", "DRS-augmented"),
  col    = c("#4E84A8", "#C1614A"),
  lwd    = 2,
  bty    = "n"
)
```

```{r fig-soc-distributions, echo = FALSE, eval = TRUE, fig.cap="Empirical distributions of depth-harmonised SOC (0–30 cm) for the reference dataset (conventional laboratory measurements only) and the augmented dataset (laboratory measurements combined with DRS-based predictions).", fig.width=6.5, fig.height=4.5}

col_wet <- "#4E84A8" # steel blue
col_spec <- "#C1614A" # muted terracotta

dens_ref <- density(soc_ref_0_30_xy$SOC_0_30, na.rm = TRUE)
dens_aug <- density(soc_aug_0_30_xy$SOC_0_30, bw = dens_ref$bw, na.rm = TRUE)

df <- rbind(
  data.frame(x = dens_ref$x, density = dens_ref$y, method = "Conventional lab"),
  data.frame(x = dens_aug$x, density = dens_aug$y, method = "DRS-augmented")
)

means_df <- data.frame(
  mean = c(wet_mean, spec_mean),
  sd = c(wet_sd, spec_sd),
  method = c("Conventional lab", "DRS-augmented")
)

ymax <- max(df$density)

ggplot(df, aes(x = x, y = density, colour = method, fill = method)) +
  geom_area(alpha = 0.20, position = "identity") +
  geom_line(linewidth = 1.0) +
  scale_colour_manual(
    values = c("Conventional lab" = col_wet, "DRS-augmented" = col_spec)
  ) +
  scale_fill_manual(
    values = c("Conventional lab" = col_wet, "DRS-augmented" = col_spec)
  ) +
  labs(
    x = "SOC, %", 
    y = "Probability density", 
    colour = NULL, 
    fill = NULL
  ) +
  
  theme_classic(base_size = 13) +
  theme(
    legend.position = "top",
    legend.key.size = unit(0.9, "lines"),
    legend.text = element_text(size = 11),
    axis.line = element_line(colour = "grey40"),
    panel.grid.major.y = element_line(colour = "grey92", linewidth = 0.4), 
    panel.border = element_rect(colour = "grey40", fill = NA, linewidth = 0.5)
  )
```


Note that the depth harmonisation step propagates horizon-level uncertainties through a variance-weighted formula, which reduces the aggregated standard deviation for profiles with multiple horizons as a mathematical artefact rather than a reflection of improved data quality. To avoid this distortion, the laboratory uncertainty is reassigned directly at the profile level before weighting:

```{r eval=TRUE}
# Based on Stevens et al. (2013), SEL ≈ 2 g C kg-1 for mineral soils
soc_ref_0_30_xy$SOC_sd_0_30 <- 0.15
soc_aug_0_30_xy$SOC_sd_0_30[!is.na(soc_aug_0_30_xy$SOC_0_30) & 
  soc_aug_0_30_xy$ProfID %in% dat_augmented$ProfID[ks_samples$model]] <- 0.15
```



## Digital soil mapping 

### Preparation of the covariates

At this stage, the environmental covariates are prepared exactly as in the training material, so the same workflow can be followed here. The objective is to align the soil observations with the raster covariate stack and extract the values of all environmental predictors for each sampling location. These predictors will later be used in the modeling step.

The procedure involves four main operations:

- loading the raster stack of environmental covariates,

- converting the soil profile summaries into a spatial point object,

- reprojecting the points so that they match the coordinate reference system of the covariates,

- extracting raster values at each soil observation location and attaching them to the point data.

This produces a spatial dataset in which each soil observation is associated with the full set of environmental predictors.

```{r message=FALSE, warning=FALSE, eval=TRUE, results = 'hide'}
library(terra)
# Load environmental covariates
covs <- rast(
  "01_data/module1/training_data/data/Environmental_Covariates_250m_KANSAS.tif"
)
cov_names <- names(covs)

# Reference dataset (laboratory measurements only)
soil_pts_ref <- vect(
  soc_ref_0_30_xy,
  geom = c("Long_Site.x", "Lat_Site.x"),
  crs = "EPSG:4326"
)
soil_pts_ref <- project(soil_pts_ref, covs)
soil_pts_ref <- cbind(soil_pts_ref, extract(covs, soil_pts_ref)[, -1])

# Augmented dataset (laboratory + DRS-based predictions)
soil_pts_aug <- vect(
  soc_aug_0_30_xy,
  geom = c("Long_Site.x", "Lat_Site.x"),
  crs = "EPSG:4326"
)
soil_pts_aug <- project(soil_pts_aug, covs)
soil_pts_aug <- cbind(soil_pts_aug, extract(covs, soil_pts_aug)[, -1])

# Optional check: plot sample locations
plot(soil_pts_ref, cex = 0.7, col = "red")
```

### DSM model training of SOC with DRS-augmented data

In this section, we demonstrate how DRS predictions can be integrated into DSM modeling. The idea is that each observation, whether derived from laboratory analysis or DRS, is associated with a sample-specific uncertainty estimate. This uncertainty can be used to control the relative influence of each observation during model training. A practical way to do this is through case weighting. Observations with lower uncertainty are assigned higher statistical weight, while observations with higher uncertainty contribute less to the fitted model. 

In the following example, we again use _quantile regression forests_, this time to map SOC following a workflow similar to the previous DSM examples. The workflow includes deriving uncertainty-based weights, fitting models with and without spectroscopic data, and comparing the resulting spatial predictions.

#### Deriving observation weights

Weights are derived from the uncertainty estimates obtained in the previous section. A simple and commonly used formulation assigns each observation a weight equal to the inverse of its variance:

$$w_i = \frac{1}{\sigma_i^2}$$

Observations with smaller uncertainty receive larger weights and therefore contribute more to model calibration. Weights are computed separately for each dataset:

- for the reference dataset, $\sigma_i$ is the depth-harmonized laboratory analytical uncertainty;

- for the augmented dataset, $\sigma_i$ is either the laboratory analytical uncertainty or the DRS prediction uncertainty, depending on the source of each observation.

Weights are then normalized by dividing by the maximum weight within each dataset, so that the most precise observation receives a weight of one and all others are expressed relative to it:

$$w_i^{\text{norm}} = \frac{w_i}{\max(w)}$$

```{r dh-weights, message=FALSE, warning=FALSE, eval=TRUE, results='hide'}
# Compute inverse-variance weights
soil_pts_ref$weight <- 1 / soil_pts_ref$SOC_sd_0_30^2
soil_pts_aug$weight <- 1 / soil_pts_aug$SOC_sd_0_30^2

# Normalise by the maximum weight within each dataset
soil_pts_ref$weight <- soil_pts_ref$weight / max(soil_pts_ref$weight, na.rm = TRUE)
soil_pts_aug$weight <- soil_pts_aug$weight / max(soil_pts_aug$weight, na.rm = TRUE)

summary(soil_pts_ref$weight)
summary(soil_pts_aug$weight)
```



#### Baseline DSM model using laboratory data

Before integrating DRS-derived information, we first fit a baseline DSM model using laboratory SOC observations only. Environmental covariates representing soil-forming factors are used as predictors in a random forest model. This provides a reference spatial prediction of SOC across the study area, against which models incorporating spectroscopic data can be compared. The resulting baseline SOC map is shown in Figure \@ref(fig:fig-mod-baseline) and was generated using the following code:

```{r fig-mod-baseline, message=FALSE, warning=FALSE, eval=TRUE, fig.cap = "Baseline DSM prediction of SOC across the study area using laboratory observations only. Environmental covariates representing soil-forming factors were used as predictors in a random forest model."}
library(ranger)

# Prepare modeling table
soil_df_ref <- as.data.frame(soil_pts_ref, geom = "XY")[ ,
  c("x", "y", "SOC_0_30", "SOC_sd_0_30", "weight", cov_names)
]
soil_df_ref <- soil_df_ref[!is.na(soil_df_ref$SOC_0_30), ]

# Fit baseline random forest model using laboratory SOC
form <- as.formula(
  paste("SOC_0_30 ~", paste(cov_names, collapse = " + "))
)

mod_ref <- ranger(
  formula = form,
  data = soil_df_ref,
  case.weights = soil_df_ref$weight,
  replace = FALSE,
  sample.fraction = 0.632,
  num.trees = 500,
  seed = 201909
)

# Predict SOC across the covariate raster stack
soc_map_ref <- predict(covs, mod_ref, na.rm = TRUE)

# Plot baseline SOC map
plot(soc_map_ref, main = "Baseline SOC map (laboratory observations only)", range = c(0, 4))
```



#### DSM model using DRS-augmented data

The augmented dataset combines laboratory SOC measurements with DRS-based predictions, increasing the number of observations available for model calibration. The modeling procedure follows the same structure as the baseline, with one difference: observation weights derived from the unified `SOC_sd` column are passed to the model, so that laboratory measurements and DRS predictions contribute proportionally to their respective uncertainties. The resulting SOC map is shown in Figure \@ref(fig:fig-mod-augmented) and was generated using the following code:

```{r fig-mod-augmented, message=FALSE, warning=FALSE, eval=TRUE, fig.cap = "DSM of SOC using the DRS-augmented data. It uses the same the same environmental covariates as for the baseline DSM."}
# Prepare modeling table
soil_df_aug <- as.data.frame(soil_pts_aug, geom = "XY")[,
  c("x", "y", "SOC_0_30", "SOC_sd_0_30", "weight", cov_names)
]

# Remove profiles with missing SOC
soil_df_aug <- soil_df_aug[!is.na(soil_df_aug$SOC_0_30), ]

# Fit random forest model using the augmented dataset
mod_aug <- ranger(
  formula = form,
  data = soil_df_aug,
  case.weights = soil_df_aug$weight,
  replace = FALSE,
  sample.fraction = 0.632,
  num.trees = 500,
  seed = 201909
)

# Predict SOC across the covariate raster stack
soc_map_aug <- predict(covs, mod_aug, na.rm = TRUE)

# Plot augmented SOC map
plot(soc_map_aug, main = "SOC map (DRS-augmented dataset)", range = c(0, 4))
```
### Model validation

The predictive performance of both DSM models is evaluated using cross-validation. Because the augmented dataset combines laboratory measurements and DRS-derived predictions, validation must account for observation-specific uncertainty. Treating all observations as equally reliable would be inconsistent with the weighting strategy applied during model calibration. Validation statistics are therefore computed using the same inverse-variance weights, so that precise observations contribute more to the performance assessment than uncertain ones.

#### Weighted performance metrics

Three weighted metrics are used. Each extends a standard DSM validation statistic by incorporating observation weights $w_i$.

The weighted mean error ($\text{ME}_w$) measures signed prediction bias:

$$\text{ME}_w = \frac{\sum_{i=1}^{n} w_i \left( \hat{z}(s_i) - z(s_i) \right)}{\sum_{i=1}^{n} w_i}$$

The weighted root mean squared error ($\text{RMSE}_w$) quantifies overall prediction error, penalising large deviations more than small ones:

$$\text{RMSE}_w = \sqrt{\frac{\sum_{i=1}^{n} w_i \left( \hat{z}(s_i) - z(s_i) \right)^2}{\sum_{i=1}^{n} w_i}}$$

The weighted Nash–Sutcliffe efficiency ($\text{NSE}_w$) evaluates predictive skill relative to the weighted mean of observed values $\bar{z}_w = \sum_i w_i z(s_i) / \sum_i w_i$:

$$\text{NSE}_w = 1 - \frac{\sum_{i=1}^{n} w_i \left( \hat{z}(s_i) - z(s_i) \right)^2}{\sum_{i=1}^{n} w_i \left( z(s_i) - \bar{z}_w \right)^2}$$

Values of $\text{NSE}_w$ close to 1 indicate good predictive performance; negative values indicate the weighted mean is a better predictor than the model.

```{r val-functions, message=FALSE, warning=FALSE}
wME <- function(obs, pred, w) {
  sum(w * (pred - obs), na.rm = TRUE) / sum(w, na.rm = TRUE)
}

wRMSE <- function(obs, pred, w) {
  sqrt(sum(w * (pred - obs)^2, na.rm = TRUE) / sum(w, na.rm = TRUE))
}

wNSE <- function(obs, pred, w) {
  zbar_w <- sum(w * obs, na.rm = TRUE) / sum(w, na.rm = TRUE)
  num <- sum(w * (pred - obs)^2, na.rm = TRUE)
  den <- sum(w * (obs - zbar_w)^2, na.rm = TRUE)
  1 - num / den
}
```

#### Cross-validating the models

Model performance is assessed using 10-fold cross-validation, applied separately to the reference and augmented datasets. In each fold, the model is trained on the remaining observations and used to predict SOC for the withheld fold. Weighted performance metrics are then computed using the same inverse-variance weights used during model calibration, so that the validation reflects the same data quality assumptions as the fitted models.

```{r val-cv, message=FALSE, warning=FALSE, eval=TRUE, results='hide'}
run_cv <- function(data, response, predictors, nfolds = 10, seed = 201909) {
  set.seed(seed)
  n <- nrow(data)
  fold_id <- sample(rep(seq_len(nfolds), length.out = n))
  cv_pred <- rep(NA_real_, n)
  form <- as.formula(paste(response, "~", paste(predictors, collapse = " + ")))
  
  for (fold in seq_len(nfolds)) {
    train <- data[fold_id != fold, ]
    test <- data[fold_id == fold, ]
    mod <- ranger(
      formula = form,
      data = train,
      case.weights = train$weight,
      replace = FALSE,
      sample.fraction = 0.632,
      seed = seed
    )
    cv_pred[fold_id == fold] <- predict(mod, data = test)$predictions
  }
  
  data.frame(
    obs = data[[response]],
    pred = cv_pred,
    weight = data$weight
  )
}

cv_ref <- run_cv(soil_df_ref, response = "SOC_0_30", predictors = cov_names)
cv_aug <- run_cv(soil_df_aug, response = "SOC_0_30", predictors = cov_names)
```

The weighted performance metrics are computed for each model and assembled into a comparison table:

```{r val-metrics, message=FALSE, warning=FALSE, eval=TRUE}
compute_metrics <- function(cv) {
  data.frame(
    wME = round(wME(cv$obs,   cv$pred, cv$weight), 3),
    wRMSE = round(wRMSE(cv$obs, cv$pred, cv$weight), 3),
    wNSE = round(wNSE(cv$obs,  cv$pred, cv$weight), 3)
  )
}

metrics <- rbind(
  data.frame(Model = "Reference (lab only)",  compute_metrics(cv_ref)),
  data.frame(Model = "Augmented (lab + DRS)", compute_metrics(cv_aug))
)

metrics
```


### Comparing maps produced with different proportions of spectroscopic data

After model validation, spatial prediction maps can be generated. Maps derived from laboratory-only models can be compared with maps obtained from models that incorporate spectroscopic predictions. This visual comparison helps to identify changes in spatial patterns, prediction smoothness, and local variability.

Differences between maps may reflect both the increased data density and the influence of uncertainty-based weighting. Such comparisons are an important step in understanding the practical impact of spectroscopic data supplementation in DSM applications.

To visualise the effect of increasing the proportion of spectroscopic observations, several mixed-data scenarios can be fitted and mapped. In the example below, four proportions are considered: 10\%, 25\%, 50\%, and 75\% spectroscopic data. For each scenario, a weighted random forest is fitted, a SOC map is predicted, and the result is compared with the baseline map derived from laboratory observations only.

The final figure is organised in two rows:
* the top row shows the SOC maps for the four mixed scenarios, with sampling points coloured by source,
* the bottom row shows the difference between each mixed map and the baseline laboratory-only map.

```{r fig-mixed-maps-build, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE, fig.cap="SOC maps produced with increasing proportions of spectroscopic observations (top row) and their differences from the baseline laboratory-only SOC map (bottom row)."}
# -----------------------------------------------------------
# Sensitivity analysis: mapping SOC using different proportions
# of spectroscopic observations
# -----------------------------------------------------------

library(terra)
library(ranger)
library(ggplot2)
library(ggpubr)

set.seed(123)

# Proportions of spectroscopic observations to evaluate
props <- c(0.10, 0.25, 0.50, 0.75)

# Storage objects for maps and differences
map_plots  <- vector("list", length(props))
diff_plots <- vector("list", length(props))

# Temporary storage to harmonise colour scales across panels
map_vals   <- vector("list", length(props))
diff_vals  <- vector("list", length(props))

# -----------------------------------------------------------
# Loop over spectroscopic data proportions
# -----------------------------------------------------------

for (i in seq_along(props)) {
  
  p <- props[i]
  n <- nrow(soil_df)
  
  # Randomly select observations replaced by spectroscopic SOC
  idx_rf <- sample(seq_len(n), size = round(p * n))
  
  # Create mixed SOC response variable
  soil_df$SOC_mix <- soil_df$SOC
  soil_df$SOC_mix[idx_rf] <- soil_df$SOC_predRF_0_30[idx_rf]
  
  # Assign uncertainty-based weights
  soil_df$w_mix <- soil_df$w_LAB_raw
  soil_df$w_mix[idx_rf] <- soil_df$w_RF_raw[idx_rf]
  
  # Track data source for plotting
  soil_df$source <- "LAB"
  soil_df$source[idx_rf] <- "RF"
  
  # Keep only valid rows for modelling
  vars_needed <- c("SOC_mix", "w_mix", cov_names)
  
  soil_df2 <- soil_df[
    complete.cases(soil_df[, vars_needed]) &
      is.finite(soil_df$w_mix) &
      soil_df$w_mix > 0,
  ]
  
  # -----------------------------------------------------------
  # Fit weighted random forest model
  # -----------------------------------------------------------
  
  mod_mix <- ranger(
    formula = as.formula(
      paste("SOC_mix ~", paste(cov_names, collapse = " + "))
    ),
    data = soil_df2,
    case.weights = soil_df2$w_mix,
    num.trees = 50
  )
  
  # Predict spatial SOC map
  soc_map_mix <- predict(covs, mod_mix, na.rm = TRUE)
  
  # Compute difference from baseline laboratory-only map
  diff_map <- soc_map_mix - soc_map
  
  # Convert rasters to data frames for ggplot visualisation
  df_map  <- as.data.frame(soc_map_mix, xy = TRUE, na.rm = TRUE)
  df_diff <- as.data.frame(diff_map,    xy = TRUE, na.rm = TRUE)
  
  names(df_map)[3]  <- "value"
  names(df_diff)[3] <- "value"
  
  # Store values to define common colour scales later
  map_vals[[i]]  <- df_map$value
  diff_vals[[i]] <- df_diff$value
  
  # -----------------------------------------------------------
  # Build map panels
  # -----------------------------------------------------------
  
  map_plots[[i]] <- ggplot() +
    geom_raster(data = df_map, aes(x = x, y = y, fill = value)) +
    geom_point(
      data = soil_df2,
      aes(x = x, y = y, color = source),
      size = 0.6
    ) +
    scale_color_manual(values = c(LAB = "red", RF = "blue")) +
    coord_equal() +
    labs(
      title = paste0(round(100 * p), "% spectro"),
      fill  = "SOC",
      color = NULL
    ) +
    theme_minimal() +
    theme(
      axis.title = element_blank(),
      axis.text  = element_blank(),
      axis.ticks = element_blank(),
      panel.grid = element_blank()
    )
  
  diff_plots[[i]] <- ggplot() +
    geom_raster(data = df_diff, aes(x = x, y = y, fill = value)) +
    coord_equal() +
    labs(
      title = paste0("Difference ", round(100 * p), "%"),
      fill  = "Diff"
    ) +
    theme_minimal() +
    theme(
      axis.title = element_blank(),
      axis.text  = element_blank(),
      axis.ticks = element_blank(),
      panel.grid = element_blank()
    )
}

# -----------------------------------------------------------
# Define common colour scales across all panels
# -----------------------------------------------------------

map_range  <- range(unlist(map_vals),  na.rm = TRUE)
diff_range <- range(unlist(diff_vals), na.rm = TRUE)
diff_lim   <- max(abs(diff_range))

for (i in seq_along(props)) {
  
  map_plots[[i]] <- map_plots[[i]] +
    scale_fill_viridis_c(limits = map_range)
  
  diff_plots[[i]] <- diff_plots[[i]] +
    scale_fill_gradient2(
      limits   = c(-diff_lim, diff_lim),
      midpoint = 0
    )
}

# -----------------------------------------------------------
# Arrange final figure layout
# -----------------------------------------------------------

row1 <- ggarrange(
  plotlist = map_plots,
  ncol = 4, nrow = 1,
  common.legend = TRUE,
  legend = "top"
)

row2 <- ggarrange(
  plotlist = diff_plots,
  ncol = 4, nrow = 1,
  common.legend = TRUE,
  legend = "bottom"
)

final_fig <- ggarrange(row1, row2, ncol = 1, nrow = 2)
final_fig
```


### Testing different ratios of spectroscopic and laboratory data

The influence of spectroscopic information on DSM can be further explored by systematically varying the proportion of spectroscopically inferred observations used during model calibration. Such a sensitivity analysis provides insight into how increasing reliance on spectroscopy may affect both the spatial pattern of predictions and the overall predictive performance of the model.

In this exercise, several scenarios are evaluated, ranging from situations where only a small fraction of the calibration dataset is derived from spectroscopy to cases where spectroscopic observations constitute a substantial component of the available information.
For each scenario, a weighted DSM model is fitted using uncertainty-based case weights, and spatial prediction maps are generated.

Difference maps relative to the laboratory-only baseline allow visual inspection of how predicted SOC distributions evolve as the balance between laboratory and spectroscopic data changes.

```{r,fig-sensitivity-metrics, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE, fig.cap="Sensitivity of weighted validation metrics to the proportion of spectroscopic observations. Points show individual repetitions, lines show the mean across repetitions, and ribbons represent one standard deviation."}
# -----------------------------------------------------------
# Sensitivity analysis: effect of increasing the proportion
# of spectroscopic observations on weighted model performance
# -----------------------------------------------------------

library(ranger)
library(ggplot2)
library(ggpubr)

set.seed(123)

# Number of observations and cross-validation settings
n <- nrow(soil_df)
k <- 10
n_rep <- 25

# Proportions of spectroscopic observations to test
perc_vec <- seq(1, 99, by = 5)

# Model formula
form <- as.formula(
  paste("SOC_mix ~", paste(cov_names, collapse = " + "))
)

# Storage table for all repetitions and proportions
results <- data.frame()

# -----------------------------------------------------------
# Loop over spectroscopic data proportions
# -----------------------------------------------------------

for (perc in perc_vec) {
  
  # Repeat each proportion several times
  # to account for randomness in the selected spectroscopic points
  for (rep in 1:n_rep) {
    
    # Randomly select the proportion of observations replaced
    # by spectroscopic SOC predictions
    idx_rf <- sample(seq_len(n), size = round((perc / 100) * n))
    
    # Create mixed SOC response
    soil_df$SOC_mix <- soil_df$SOC
    soil_df$SOC_mix[idx_rf] <- soil_df$SOC_predRF_0_30[idx_rf]
    
    # Assign the corresponding uncertainty-based weights
    soil_df$w_mix <- soil_df$w_LAB_raw
    soil_df$w_mix[idx_rf] <- soil_df$w_RF_raw[idx_rf]
    
    # Optional source indicator
    soil_df$source <- "LAB"
    soil_df$source[idx_rf] <- "RF"
    
    # Keep only complete and valid observations
    vars_needed <- c("SOC_mix", "w_mix", cov_names)
    
    soil_df2 <- soil_df[
      complete.cases(soil_df[, vars_needed]) &
        is.finite(soil_df$w_mix) &
        soil_df$w_mix > 0,
    ]
    
    # -----------------------------------------------------------
    # Create 10-fold cross-validation splits
    # -----------------------------------------------------------
    
    fold_id <- sample(rep(1:k, length.out = nrow(soil_df2)))
    cv_pred <- rep(NA_real_, nrow(soil_df2))
    
    # -----------------------------------------------------------
    # Run 10-fold cross-validation
    # -----------------------------------------------------------
    
    for (fold in 1:k) {
      
      train_data <- soil_df2[fold_id != fold, ]
      test_data  <- soil_df2[fold_id == fold, ]
      
      mod_cv <- ranger(
        formula = form,
        data = train_data,
        case.weights = train_data$w_mix,
        num.trees = 50,
        importance = "none",
        verbose = FALSE
      )
      
      cv_pred[fold_id == fold] <- predict(mod_cv, data = test_data)$predictions
    }
    
    # -----------------------------------------------------------
    # Compute weighted validation metrics for this repetition
    # -----------------------------------------------------------
    
    cv_results <- data.frame(
      obs  = soil_df2$SOC_mix,
      pred = cv_pred,
      w    = soil_df2$w_mix
    )
    
    results <- rbind(
      results,
      data.frame(
        perc_spectro = perc,
        repetition   = rep,
        wME   = wME(cv_results$obs, cv_results$pred, cv_results$w),
        wRMSE = wRMSE(cv_results$obs, cv_results$pred, cv_results$w),
        wNSE  = wNSE(cv_results$obs, cv_results$pred, cv_results$w)
      )
    )
  }
}

# -----------------------------------------------------------
# Summarise results across repetitions
# -----------------------------------------------------------

sum_mean <- aggregate(
  cbind(wME, wRMSE, wNSE) ~ perc_spectro,
  data = results,
  FUN = mean
)

sum_sd <- aggregate(
  cbind(wME, wRMSE, wNSE) ~ perc_spectro,
  data = results,
  FUN = sd
)

results_plot <- merge(
  sum_mean,
  sum_sd,
  by = "perc_spectro",
  suffixes = c("_mean", "_sd")
)

# -----------------------------------------------------------
# Plot weighted performance metrics
# Points show individual repetitions
# Line shows the mean across repetitions
# Ribbon shows ± 1 standard deviation
# -----------------------------------------------------------

p1 <- ggplot() +
  geom_point(
    data = results,
    aes(x = perc_spectro, y = wME),
    alpha = 0.5,
    size = 1
  ) +
  geom_ribbon(
    data = results_plot,
    aes(
      x = perc_spectro,
      ymin = wME_mean - wME_sd,
      ymax = wME_mean + wME_sd
    ),
    alpha = 0.2
  ) +
  geom_line(
    data = results_plot,
    aes(x = perc_spectro, y = wME_mean),
    linewidth = 0.8
  ) +
  labs(
    x = "% spectro data",
    y = "wME"
  ) +
  theme_minimal()

p2 <- ggplot() +
  geom_point(
    data = results,
    aes(x = perc_spectro, y = wRMSE),
    alpha = 0.5,
    size = 1
  ) +
  geom_ribbon(
    data = results_plot,
    aes(
      x = perc_spectro,
      ymin = wRMSE_mean - wRMSE_sd,
      ymax = wRMSE_mean + wRMSE_sd
    ),
    alpha = 0.2
  ) +
  geom_line(
    data = results_plot,
    aes(x = perc_spectro, y = wRMSE_mean),
    linewidth = 0.8
  ) +
  labs(
    x = "% spectro data",
    y = "wRMSE"
  ) +
  theme_minimal()

p3 <- ggplot() +
  geom_point(
    data = results,
    aes(x = perc_spectro, y = wNSE),
    alpha = 0.5,
    size = 1
  ) +
  geom_ribbon(
    data = results_plot,
    aes(
      x = perc_spectro,
      ymin = wNSE_mean - wNSE_sd,
      ymax = wNSE_mean + wNSE_sd
    ),
    alpha = 0.2
  ) +
  geom_line(
    data = results_plot,
    aes(x = perc_spectro, y = wNSE_mean),
    linewidth = 0.8
  ) +
  labs(
    x = "% spectro data",
    y = "wNSE"
  ) +
  theme_minimal()

# Final combined figure
perf_fig <- ggarrange(p1, p2, p3, ncol = 1, nrow = 3)
perf_fig

```


## Summary 

In this practical exercise we demonstrated how soil spectroscopic information can be integrated into a digital soil mapping workflow. Mid-infrared spectral data were first preprocessed and used to build a calibration model that provided, for each soil sample, both a predicted SOC value and a sample-specific estimate of prediction uncertainty. These spectroscopic predictions were then harmonised to a standard depth interval and combined with laboratory observations. Laboratory analytical uncertainty was also approximated so that both data sources could be treated consistently.

A key step of the workflow was the use of uncertainty-based case weights during spatial model training. By assigning weights proportional to the inverse of observation variance, laboratory data and spectroscopic predictions contributed to model calibration according to their relative reliability. Model validation was also performed using weighted performance metrics, ensuring that predictive skill was evaluated in a way that reflects heterogeneous data quality. Spatial prediction maps illustrated how supplementing laboratory data with spectroscopy can modify spatial patterns and potentially improve prediction stability, depending on the proportion and uncertainty of the additional data.

Although the example focused on SOC mapping with quantile regression forests, the same principles can be applied to other soil properties and modelling approaches. Only minor adaptations are required, such as changing the calibration model, the spatial prediction algorithm, or the method used to propagate uncertainty. More broadly, the workflow is a practical framework for integrating emerging proximal sensing data sources into DSM while maintaining a transparent treatment of data quality and prediction uncertainty.
